## Appendix Constitutional Changes

{{< pagebreak >}}

```{r}
#| output: asis 
#| echo: false 
#| warning: false 
#| label: tbl-populistchanges 
#| tbl-cap: Constitutional changes under populist governments (weighted populism score > 0.5)  

# Read the .tex file content
tex_content <- readLines("results/tables/populistchanges.tex")  
cat("\\scriptsize")
# Print the content as raw LaTeX 
cat(tex_content, sep="\n")
```

\blandscape

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-resultschange_EU
#| tbl-cap: Regression results for likelihood of constitutional change

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchangelikelihood_EU_evnt.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-executive
#| tbl-cap: Regression models for changes in rights of the executive

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchangelikelihood_diff_executive.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")


```

\elandscape

![Average marginal effect of constitutional change condititoned by government weighted populism score and ideology (based on table @tbl-results).](results/graphs/change_effect_appendix.pdf){#fig-interactionapp fig-align="center" fig-pos="H" fig-scap="Average marginal effect of constitutional change condititoned by government weighted populism score and ideology"}

![Coefficients of interaction effect of jackknifed models for @tbl-results, model 3.](results/graphs/jackknife_change.pdf){#fig-jackknifechange fig-pos="H" fig-scap="Coefficients of interaction effect of jackknifed models" fig-align="center" width="90%"}

![Predicted change in rights for the executive conditioned by ideology and populism based on @tbl-executive, model 4.](results/graphs/change_in_executive.pdf){#fig-executive fig-pos="H" fig-scap="Predicted change in rights for the executive conditioned by ideology and populism score" fig-align="center" width="90%"}

```{=tex}
\blandscape
\renewcommand{\arraystretch}{0.5}
\setlength{\tabcolsep}{2pt}
```
```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-libdem
#| tbl-cap: Full regression models on liberal democracy

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_v2x_libdem.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{=tex}
\renewcommand{\arraystretch}{0.5}
\setlength{\tabcolsep}{2pt}
```
```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-polyarchy
#| tbl-cap: Full regression models on polyarchy

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_v2x_polyarchy.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{=tex}
\renewcommand{\arraystretch}{0.5}
\setlength{\tabcolsep}{2pt}
```
```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-partip
#| tbl-cap: Full regression models on participation

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_v2x_partip.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{=tex}
\renewcommand{\arraystretch}{0.5}
\setlength{\tabcolsep}{2pt}
```
```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-egaldem
#| tbl-cap: Full regression models on polyarchy

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_v2x_egaldem.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{=tex}
\renewcommand{\arraystretch}{0.5}
\setlength{\tabcolsep}{2pt}
```
```{r}
#| output: asis
#| warning: false
#| echo: false
#| label: tbl-cspart
#| tbl-cap: Full regression models on civil society

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_v2x_cspart.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

\elandscape

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-libdem_sep
#| tbl-cap: Separate regression models on liberal democracy for Europe and Latin America and surplus model

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_alternative_v2x_libdem.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-polyarchy_sep
#| tbl-cap: Separate regression models on polyarchy for Europe and Latin America  and surplus model

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_alternative_v2x_polyarchy.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-partip_sep
#| tbl-cap: Separate regression models on participation for Europe and Latin America  and surplus model

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_alternative_v2x_partip.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-egaldem_sep
#| tbl-cap: Separate regression models on egalitarian democracy for Europe and Latin America and surplus model

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_alternative_v2x_egaldem.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

```{r}
#| output: asis
#| warning: false
#| echo: false
#| label: tbl-cspart_sep
#| tbl-cap: Separate regression models on civil society for Europe and Latin America and surplus model

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_alternative_v2x_cspart.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

![Average marginal effect of constitutional change condititoned by government weighted populism score and ideology (GAL-TAN) for Europe, based @tbl-libdem_sep to @tbl-cspart_sep, model 3.](results/graphs/constchange_eu.pdf){#fig-eu fig-pos="H" fig-scap="Average marginal effect of constitutional change condititoned by government weighted populism score and ideology (GAL-TAN)" fig-align="center"}

![Average marginal effect of constitutional change condititoned by government weighted populism score and ideology for Latin America, based @tbl-libdem_sep to @tbl-cspart_sep, model 2.](results/graphs/constchange_la.pdf){#fig-la fig-pos="H" fig-scap="Average marginal effect of constitutional change condititoned by government weighted populism score and ideology for Latin America" fig-align="center"}

```{=tex}
\blandscape
\renewcommand{\arraystretch}{0.5}
```
```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-dynamic
#| tbl-cap: Dynamic regression models 

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_dynamic.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")


```

\elandscape

![Average marginal effect of constitutional change conditioned by government weighted populism score and lag of democratic quality before, based @tbl-dynamic.](results/graphs/constchange_dynamic.pdf){#fig-dynamic fig-pos="H" fig-scap="Average marginal effect of constitutional change conditioned by government weighted populism score and lag of democratic quality" fig-align="center"}

\blandscape

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-ruth
#| tbl-cap: Regression models with Ruth-Lovell & Grahn populism coding

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_ruth.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

\renewcommand{\arraystretch}{0.5}

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-leadlibdem
#| tbl-cap: Regression models on liberal democracy for different leads

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_lead_v2x_libdem.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-leadpartip
#| tbl-cap: Regression models on participation for different leads

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_lead_v2x_partip.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-leadpolyarchy
#| tbl-cap: Regression models on polyarchy for different leads

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_lead_v2x_polyarchy.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-leadcspart
#| tbl-cap: Regression models on civil society for different leads

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_lead_v2x_cspart.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-leadegaldem
#| tbl-cap: Regression models on egalitarian democracy for different leads

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_lead_v2x_egaldem.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

\elandscape

\blandscape

![Coefficients of triple-Interaction effect in jackknife-model.](results/graphs/jackknife_plots.pdf){#fig-jackknife1 fig-align="center" fig-pos="H"}

![Coefficients of triple-interaction effect in jackknife-model.](results/graphs/jackknife_plots2.pdf){#fig-jackknife2 fig-align="center" fig-pos="H"}

\elandscape

# References

\normalsize
