---
title: "The Relationship between Democratic and Constitutional Regression under Populist Governments"
subtitle: "An Empirical Analysis"
format: pdf
editor: visual
documentclass: article
author:
  - name: Jasmin Sarah König^[University of Hamburg. Jasmin wurde durch die Deutsche Forschungsgemeinschaft (DFG) – GRK 2503 gefördert. Address for Correspondence jasmin.sarah.koenig@uni-hamburg.de]
    email: jasmin.sarah.koenig@uni-hamburg.de
    affiliation: University of Hamburg
  - name: Tilko Swalve^[Leibniz University Hannover]
bibliography: citation/lit.bib
csl: citation/ejpr.csl
classoption: abstract
abstract: |
  Populism and liberal democracy are – at least in parts – in conflict. Researchers have discussed a possible relationship between populist parties in government and democratic regression. In countries with sweeping populist majorities - such as Hungary - populists in power have undermined democratic institutions through constitutional changes. Our paper analyzes whether this mechanism is systematically applied by populists in government in Europe and Latin America. Using V-Dem and V-Party data, we investigate whether constitutional change leads to democratic regression more often under populist governments. The results from our multi-level model show that the relationship between populism and constitutional regression is ambiguous. 
include-in-header: 
  text:
    \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    \renewcommand{\abstractname}{Abstract}
    \usepackage{booktabs}
    \usepackage{siunitx}
    \newcolumntype{d}{S[
      input-open-uncertainty=,
      input-close-uncertainty=,
      parse-numbers = false,
      table-align-text-pre=false,
      table-align-text-post=false
      ]}
---

```{r}
#| include: false
options(tinytex.verbose = TRUE)

```

<br>

Accepted for [Peter Niesen (eds): Zur Diagnose demokratischer Regression. Leviathan, Sonderband 40. Im Erscheinen](https://www.nomos-shop.de/nomos/titel/zur-diagnose-demokratischer-regression-id-105862/)

{{< pagebreak >}}

# 1. Introduction

Populism and liberal democracy are - at least in part - incompatible. The anti-pluralist elements of the populist ideology and the liberal democratic imperative to enforce the will of the majority are at odds with the preservation of minority rights and the influence of non-majoritarian institutions in liberal democracies. This is why many authors link the increasingly common diagnosis of democratic regression to populism as a driving force [@Schafer.2021; @Ginsburg.2018; @Diamond.2021]. However, democratic regression or progress happens not only with regard to liberal democratic institutions but also on other democratic dimensions, such as participation or egalitarianism. To understand the relationship between populism and democratic regression in more detail, we analyze whether constitutional changes implemented by populist parties in government lead to decreases of democratic quality on different dimensions. We further test whether this effect differs between right-wing and left-wing populist parties in power.

The concept of democratic regression and the mechanisms behind it are controversial in the literature[@Bermeo.2016; @Karolewski.2021; @Jee.2021; @Wolkenstein.2022; @Waldner.2018]. We therefore examine in this article one particular aspect of democratic regression, constitutional regression. To undermine democracies, it often takes constitutional amendments that keep the veneer of democratic action intact [@Scheppele.2018] We understand constitutional regression to mean constitutional amendments that lead to a decline in democratic quality.

Time and again, the importance of constitutional change is discussed in the context of populists in power [@Landau.2013; @Scheppele.2018; @Huq.2018; @Muller.2017b; @Muller.2017; @Blokker.2020]. The assumption that constitutional change under populist governments leads to democratic regression more often than under mainstream governments is based on the fact that the populist ideology is in particular tension with constitutionalism. In populism, law is understood as the result of the "will of the people" and thus as political [@Blokker.2019]. This suggests that constitutional reforms happen more frequently under populist governments.

We argue that the relationship between constitutional change under populist governments and democratic quality is not as clear as the literature on democratic regression suggests [@Landau.2013; @Scheppele.2018; @Huq.2018]. While populism is in conflict with liberal democratic institutions, it can have a positive impact on participation or egalitarianism [@Mudde.2012b]. Similar to a study by @RuthLovell.2022, we analyze the effects on democratic quality on distinct democratic dimensions, liberal democracy, polyarchy, egalitarianism, civil society and participation.

Using data from the V-Dem [@Vdem.2021] and V-Party [@VParty.2022] as well as the Comparative Constitutions Project [@ComparativeConstitutions.2005], we examine what effects constitutional change under populist governments has on democratic quality. To approach these questions, we conduct a Large-N analysis of the impact of constitutional change on the quality of democracy under populist and non-populist governments. Our results show that the effect of constitutional amendments by populist parties in power differ between democratic dimensions and party ideology. We do not find any significant negative effects of constitutional changes under populists in power on democratic quality. But, left-wing populist parties have had a significant positive impact on liberal democracy, polyarchy, and egalitarianism.

Building on our large-N analysis, we discuss to case studies to highlight the difference between changes in ordinary law and constitutional law. We compare developments in Hungary and Poland over the past twenty years. In doing so, we show that a similar decline in the quality of democracy in both countries can be associated with both constitutional regression and changes in laws below the constitutional level. But, to reverse the implemented changes becomes more challenging for future governments if constitutional regression took place [@AydinCakir.2023].

# 2. Consitutional Regression and Populism

To discuss the relationship between populism and constitutional regression, we first introduce the concept of democratic regression. Based on this discussion, we show that law has a special significance in the process of democratic regression. We discuss different approaches to the meaning of law in democratic regression and describe our understanding of constitutional regression. In the next step, we use the definition of populism as a thin ideology[@Mudde.2004] to explain the relationship between populism and constitutionalism.

## 2.1 Definition of Constitutional Regression

Democratic regression refers to a decline in democratic quality. What exactly constitutes a decline in democratic quality is disputed [@Gerschewski.2021]. Scholars agree that democratic regression happens more slowly in recent years and is mostly carried out by (at least nominally) democratically elected governments. In the past, democratic systems were often undermined by coups [@Bermeo.2016]. Today, incumbents and officeholders who want to secure their power undermine democratic rules and values, and with it the democratic system [@Bermeo.2016; @Huq.2018; @Levitsky.2018; @Svolik.2015].

Going further into the details of democratic regression, we find more discussions around the term. While some authors argue that democratic regression always takes place within a democracy and thus can never describe the collapse of a democracy [@TOMINI.2018], other authors also understand a hybrid system between democracy and autocracy as a possible outcome [@Bermeo.2016 ; @Huq.2018]. In our analysis, we apply the latter understanding and also include periods in which states were already considered electoral (but not closed) autocracies in order to gain the most comprehensive understanding of the influence of populism on constitutional regression.

The core of democratic regression is usually described as executive aggrandizement: the strengthening of executive power while constraining veto players and opposition [@Bermeo.2016 ; @Ginsburg.2018; @Scheppele.2018, p. 549; @Satrio.2018, pp. 276-277; @Karolewski.2021, p. 313; @Waldner.2018, p. 95; @Khaitan.2019, p. 344; @Diamond.2021]. In the course of democratic regression, checks and balances within the political system are weakened, leaving the government subject to fewer and fewer checks and balances.

Which other mechanisms are part of democratic regression, however, varies widely by definition. Some authors speak of democratic regression only when free elections are also restricted, for example by redrawing constituencies, manipulating the media exposure of candidates, or making it more difficult for certain groups to participate in elections [@Huq.2018, p. 83; @Waldner.2018, p. 95]. Likewise, the restriction of opposition, freedom of the press, freedom of movement, freedom of association, and science and media are discussed as elements of democratic regression [@Huq.2018, p. 83; @TOMINI.2018, p. 693; @Levitsky.2018; @Diamond.2021 ]. Less frequently mentioned are a decline in citizen participation for example due to declining voter turnout or a dismantling of direct democratic elements, as well as regressions in the inclusion of disadvantaged population groups [@Waldner.2018, p. 95; @TOMINI.2018, p. 693.].

The debate on democratic regression often focuses on the weakening of liberal democratic institutions and (in-)formal norms [@Bermeo.2016; @Levitsky.2018; @Diamond.2021]. This is increasingly criticized. Authors have recently started to emphasize the importance of civil society in the concept of democratic regression [@Karolewski.2021; @Wolkenstein.2022; @Schafer.2021]. They argue that decreasing inclusion, or fewer participation mechanisms, as well as citizens' attitudes towards democracy should be considered as an aspect of democratic regression as well. Sometimes a decrease in the quality of liberal democratic institutions might even be considered as an increase in other aspects of democratic quality. A democracy can also fail to serve its purpose and lose quality if all institutions are still intact but citizens no longer want to be involved in the political process and slip into a passive behavior. Schäfer and Zürn additionally point out that representation is not always responsive and that the interests of different social groups are not always equally well represented [@Schafer.2021, pp. 96-101]. A growing lack of responsiveness and an increasing inequality in the representation of different social groups can also mean a decline in democratic quality. We account for this in our analysis by using multiple indices to measure democratic quality.

### The Importance of Legal Changes in Democratic Regression

Liberal democratic regression is closely linked to legal change. In many liberal democracies, the norms and institutions described are enshrined in constitutions, or at least in ordinary law. This gives the mechanism of (constitutional) legal change a special significance within the process of democratic regression [@Huq.2018; @Scheppele.2018].

Particularly when officeholders and incumbents aim to ensure that their state continues to appear democratic, law gains weight in the process of democratic regression [@Landau.2013]. In the case of an unconstitutional act such as a coup, it matters little whether the law is adapted. Today, however, democratic regression often takes place under the cover of democratic government. Changes in the law can be an important part of this cover to legitimize the autocratic goals of democratically elected officeholders and incumbents." [@Scheppele.2018]

The undermining of liberal democratic institutions through (constitutional) legal change has been debated in recent years under terms such as constitutional retrogression [@Huq.2018], autocratic legalism [@Scheppele.2018], and abusive constitutionalism [@Landau.2013]. The relevance of law in the course of democratic regression can be seen in the multitude of terms that have emerged in recent years to describe this process alone. Almost every article dealing with democratic regression also addresses changes in law [@Scheppele.2018; @Gerschewski.2021; @Karolewski.2021; @Huq.2018; @Khaitan.2019; @Bermeo.2016].

The term constitutional retrogression already points to the relevance of constitutions in the context of democratic regression [@Huq.2018]. At its core, the authors focus on the restriction of elections, freedom of expression and assembly, and the rule of law. By rule of law, the authors mean "stability, predictability, and integrity of law and legal institutions."[@Huq.2018, pp. 87-88] While this definition is directed at various rights, their design, and the legal system, the authors explicitly point out that constitutional amendments are only one mechanism of constitutional retrogression [@Huq.2018, p. 124.]. Huq and Ginsburg also include in their concept legal amendments that violate the character of the constitution but do not require a constitutional amendment. Scheppele pursues a similar concept with the term autocratic legalism [@Scheppele.2018]. She, too, is concerned with amendments to constitutions or simple law. However, her work is primarily devoted to restrictions on the separation of powers as well as the centralization of executive power.

The concept of abusive constitutionalism is limited to constitutional amendments or renewals aimed at undermining democracy [@Landau.2018]. The goal here is said to be -- along with the elements of democratic regression discussed earlier -- the centralization of power in the executive branch, the restriction of veto players and elections, and the curtailment of minority and fundamental rights.

The importance of the distinction between constitutional law and ordinary law can be seen in our case study analyses of Hungary and Poland: Whereas in Hungary the constitution was amended to limit the independence of the judiciary, in Poland this was possible even without this step -- and thus with a smaller parliamentary majority. We argue, however, that categorizing both types of regression together does not do justice to the situation, even if they are similar: A restructuring of liberal democratic institutions through constitutional amendments is a more permanent step. To reverse constitutional regression, many states require larger or long-term majorities in order to amend the constitution. We discuss this problem in more detail in the empirical section, using Hungary as a case study.

Therefore, our understanding of constitutional regression builds on the notion of abusive constitutionalism. In line with Landau, we include only constitutional amendments in our analysis. We thus understand constitutional regression as one or more constitutional amendments that lead to a decline in democratic quality, but include multiple democratic dimensions. Our understanding of democratic and constitutional regression thus goes beyond the institution-based definitions of Scheppele, Landau, Huq, and Ginsburg. As we show in the discussion of populism in the coming section, this distinction is important, to assess the effect of populism on constitutional regression and to address the criticism of authors who see the notion of democratic regression as too institution-based [@Karolewski.2021; @Wolkenstein.2022].

## 2.2 Populism and Constitutional Regression

The debate on democratic regression repeatedly links the development to the growing number of cases in which populist parties are involved in democratic governments [@Schafer.2021; @Ginsburg.2018; @Diamond.2021]. As noted above, looking at the tension between populism and constitutionalism raises the question of whether constitutional amendments by populist parties lead to a deterioration in the quality of democracy. We first introduce the concept of populism in this section before discussing its relationship to constitutionalism and democratic quality in the next step.

### Definition

The most commonly used definition of populism in Political Science describes it as a thin ideology [@Mudde.2004]. Populism usually emerges tied to a "denser" host ideology, for example fascism and socialism, and yet has its own ideological core. Abts and Rummens describe this core as a question of the distribution of power [@Abts.2007, p. 409]. The populist ideology is based on the majority principle, in which only the "will of the people" is to be implemented. According to the ideology, the will of the people is homogeneous and therefore clearly recognizable [@Abts.2007; @Mudde.2004]. With the goal of implementing the "true" will of the people, populism divides society into the "evil elites" and the "good people" [@Mudde.2004, p. 543]. If the alleged"will of the people" is not implemented, populists see the "evil elite" at work.

Populism and liberal democracy are thus in conflict. While populism strives for pure and unrestricted popular sovereignty as the ideal, liberal democracy is characterized by separation of powers, minority rights and veto players. In a liberal democracy, checks and balances are built in both, within and between, the powers, preventing a c entralization of power. Liberal democracy is thus always in tension between institutionalization and popular sovereignty [@Canovan.1999]. This inherent incongruence is always part of liberal democracies. Populism, however, has a clear alignment between the poles: The "will of the people" must be implemented as quickly as possible and without obstacles. Institutions and norms serve the sole purpose of supporting this process, but must never hinder it.

### Populism and Constitutionalism

Populism is based on the supremacy of the political [@Mudde.2012, p. 17; @Blokker.2019d]. Law can accordingly only express the outcome of political processes, but can never justify their restriction. The claim of a neutral law that stands above the political process is not recognized by populism. Instead, law is seen as a purely political medium. The goal of directly implementing the alleged "will of the people" leads to a special relationship between populism and constitutionalism. The constitution is supposed to reflect the will of the majority of the people and is thus not seen as a firmly established institution that is rarely changed, but as purely political [@Blokker.2020d; @Mazzoleni.2020].

Populist parties do not strive to abolish constitutions, but to re-politicize them in the sense of the alleged "will of the people."[@Mazzoleni.2020; @Muller.2017b]. In order for this to succeed, constitutions should be easy and quick to change according to the populist ideal, in order to always reflect the will of the majority [@Fabbrizi.2020]. With this understanding of the constitution, the distinction between ordinary and constitutional law is also abolished within the populist ideology [@Blokker.2020d]. If constitutional law is no longer seen as a guideline in the everyday political process, but only as a form of expression of the political, it loses its elevated and particularly safeguarded position.

### Populism and Constitutional Regression

This relationship between populism and constitutionalism raises the question of whether constitutional regression and populism are related. A vast amount of literatur has been written baout the general relationship between populism and democratic quality. Schäfer and Zürn argue that populist parties have no positive influence either in opposition or in power -- as junior, senior partners or in a single-party government. As a strong party in government they attest populists a negative effect on the quality of democracy [@Schafer.2021, pp. 171-182]. In the wider literature, however, the relationship between democracy and populism is much more controversial [@Canovan.1999; @Meny.2002b; @Mouffe.2005; @Laclau.2005; @Abts.2007; @Mudde.2012b; @KONIG.2022b]. For example, in a study of the populist FPÖ's participation in government in Austria, we did not find any above-average transgressions of constitutional limits during their time in government [@KONIG.2022b]. While many authors agree that populism and liberal democracy are at least partially in conflict, they frequently also point out that populism may well have a corrective effect on a less responsive, highly formalized democracy [@LaTorre.2020; @Mudde.2012b; @Canovan.1999]. Both @Blokker.2019d and @Bugaric.2019 refer to the variance of populist parties and argue that populist constitutional amendments do not only have negative effects on the quality of democracy. Solely a preference for frequent constitutional amendments is not a threat to democracy [@Tushnet.2021, p. 73].

The claim that society consists of a "homogeneous people" with a unified will is also reflected in populism's understanding of the constitution. While liberal democratic constitutions focus on fundamental and human rights, separation of powers, and these days also on international integration, populist parties have shifted their focus to the constitution as a whole. The understanding of the people as a homogeneous entity thus leads to the exclusion of some people from the constitution, so that it often loses its pluralistic claim [@Muller.2017b]. In Blokker's words, "Constituent power, rather than being the power of the multitude, becomes the power of the majority."[@Blokker.2020d]

Since the populist ideology centers around popular sovereignty, opportunities for participation have a special significance for populists. In particular, the strengthening of participatory elements of constitutions is repeatedly highlighted as a positive effect of populist constitutional amendments [@Blokker.2019c; @Mudde.2013b; @LaTorre.2020, p. 1464].

> **Hypothesis 1:** Constitutional Change under populist governments have a positive impact on the quality of participation.

Moreover, in Latin America, constitutional amendments stemming from populist actors have strengthened social rights [See @LaTorre.2020, p. 1464] and the inclusion of indigenous groups in the political process [@Mudde.2013b, p. 162]. Right-wing populists on the other hand are more likely to exclude groups, such as migrants, from civil society. Thus, we expect a significant interaction effect between ideology, populism and constitutional change.

> **Hypothesis 2a:** Constitutional Change under left-wing populist governments have a positive impact on the quality of civil society.
>
> **Hypothesis 2b:** Constitutional Change under populist governments have a negative impact on the quality of civil society.

A similar argument can be made with regard to the democratic dimension of egalitarianism. Left-wing populists parties often emphasize that all people need equal opportunities to participate in a democracy and to be heard [@RuthLovell.2022]. The dimension of egalitarianism also includes the idea that in order to reach equal opportunities, the state is supposed to redistribute wealth, a typical left-wing demand [@Hilgers.2013]. Thus, we expect a positive effect of constitutional change on egalitarian democracy under left-wing populists but a negative effect under right-wing populists.

> **Hypothesis 3a:** Constitutional Change under left-wing populist governments have a positive impact on the quality of egalitarianism.
>
> **Hypothesis 3b:** Constitutional Change under populist governments have a negative impact on the quality of egalitarianism.

Despite the indications of possible positive effects of constitutional changes under populist governments, most authors point out that these positive developments come at a price. Its positive attitude toward participation and popular sovereignty is countered by populism's majoritarian character. The understanding of the popular will as homogeneous often leads populist governments to weaken constitutional elements that protect social pluralism and the institutionalized separation of powers: Most often, populist constitutional amendments limit opposition rights as well as political competition and strengthen the power of the executive [@LaTorre.2020; @Mudde.2013b; @LaTorre.2018]. The notion of the people in populist parties necessarily excludes parts of society, as otherwise the idea of unrestricted homogeneity is untenable [@Albertazzi.2013; @Urbinati.2019; @LaTorre.2020, p. 1464]. The participatory elements of populist constitutional amendments are thus only suitable to allow parts of society to participate. At the same time, however, they often disenfranchise other parts by privileging the collective over individual fundamental rights." [@Scheppele.2018.]

> **Hypothesis 4:** Constitutional Change under populist governments have a negative impact on the quality of polyarchy.
>
> **Hypothesis 5:** Constitutional Change under populist governments have a negative impact on the quality of liberal democracy.

In the debate on the effect of populism on liberal democracies, authors repeatedly refer to Latin American cases in which participatory elements have been strengthened [@Mudde.2013b; @Blokker.2019c]. However, Schäfer and Zürn consider only two Latin American cases in their analysis, Venezuela and Brazil [@Schafer.2021, p. 173]. While a left-wing populist is in power in Venezuela, a right-wing populist rules in Brazil. The latter is atypical for Latin America, which has long been characterized by left-wing populism [@deLaTorre.2017]. According to the latest V-Dem report, Venezuela is one of the most autocratic states worldwide [@Boese.2022]. By not discussing an example of a left-wing populist regime with a higher quality of democracy, Schäfer and Zürn obscure the connection between democracy and populist governments in Latin America.

Overall, the case selection of the eight examples used by the authors to examine the influence of populist majority governments does not appear to be independent of their dependent variable, as only electoral autocracies and states with a sharp decline in democratic quality are considered [@Schafer.2021, p. 173]. The authors leave out states in which the effect of populism on democratic quality cannot be classified so clearly (for example, Greece or Bolivia[@Mudde.2013b; @Blokker.2019d] ). Although Schäfer and Zürn put forward the thesis that current populism is "primarily an authoritarian populism"[@Schafer.2021, p. 167], they do not examine this assumption empirically. Therefore, their study allows conclusions only for the included eight states, but not fundamentally for the relationship between populism and democracy. What drives the negative effect on democratic quality in these cases, whether it is authoritarianism or populism, remains unclear. This also applies to the question of whether the prevailing populism is actually authoritarian and what effect populism has on the quality of democracy.

The relationship between populism and democracy is more complex than the black-and-white picture often painted of the populist threat to democracy, especially when taking into account different democratic dimensions -- such as participation, inclusion, representation or non-majoritarian institutions. Many theories of populism clearly point out that populist actors have a different focus in their conception of democratic systems. While it would stand to reason that non-majoritarian institutions suffer under populist governments, this is not necessarily true for other democratic dimensions such as participation [See @Blokker.2019c; @Mudde.2013b; @LaTorre.2020, p. 72].

A further question, which unfortunately cannot find a place in this article, is which participation mechanisms are enabled in constitutional amendments [Cf @LaTorre.2018]. In order to meet the claim of general participation, from a democratic perspective, the design of the new constitution or constitutional amendments must also be inclusive. However, if the new constitution is decided top down by a populist government, this rather indicates an authoritarian development [@LaTorre.2020]. Unfortunately, no comparable data are available on the diverse processes of constitutional reforms. We therefore limit ourselves to analyzing the effect of constitutional change by populist governments on democracy quality. In the next section, we present the data used for this purpose and our empirical strategy.

# 3. Data and Empirical Strategy

If constitutional amendments come about under populist governments, the question arises whether they always harm the quality of democracy. To gain an overview of the effects of constitutional change by populist governments, we use data from the V-Dem [@vdemdata.2022], V-Party [@VParty.2022] and Comparative Constitutions Projects [@ComparativeConstitutions.2005].

To determine the populism score of a government, we draw on the populism index from the V-Party project [@VParty.2022], which ranks parties on a populism scale based on expert assessments of the rhetoric of party representatives regarding their anti-elitist attitudes and their reference to the people as a homogeneous group (v2xpa_popul). When a government consists of multiple parties, an average was taken of the populism scores of the parties involved, each weighted by their relative strength (measured by percentage of seats) within the governing coalition. This results in a populism index for governments between 0 (not populist) and 1 (populist).

We code whether a government is left- or right-wing based on the government parties' economic left-right scores [@VParty.2022]. Again, we calculate a weighted score that ranges from -3.43 (far-left) to 2.99 (far-right). Governments scoring lower than 0 are coded as left-wing, governments scoring higher than 0 are coded as right-wing. Using an economic left-right scale omits other political dimensions [@Huber.2022]. But, the scale is comparable and meaningful across continents, different to GAL-TAN scales which are often used as a second dimension and which do not show a lot of variance in Latin America [@MartinezGallardo.2023]. The GAL-TAN scale further includes aspects such as authoritarianism which are highly correlated with our dependent variable of interest. Therefore, the economic left-right scale allows the best identification for our models.

The data on constitutional changes comes from the Comparative Constitutions Project [@ComparativeConstitutions.2005]. We use the variable indicating whether a constitutional change, through amendment or replacement, occurred (0 -- no change, 1 -- change).

We measure the quality of democracy with the democracy indices from the Varieties of Democracies (V-Dem) project [@Vdem.2021]. Similar to @RuthLovell.2022, we use the different indices on participation, civil society, electoral, egalitarian, and liberal democracy to estimate the effect of changes under populist governments on different dimensions of democracy. Each index describes on a scale of 0 to 1 to what extent the ideals of the democratic dimension are implemented in the respective country-year observation.

We restrict our analysis to 57 European and Latin American countries over the period 1991-2020 (the included time-frame ranges between 12 and 29 years between countries). Country-year observations in which a regime is classified as a closed autocracy by V-Dem are excluded from our analysis, thus only the effect on already democratic states is analyzed. In total, our data panel consists of 1539country-year observations. Of these, constitutional changes take place in 561 observations. However, constitutional events by governments with a high populism score of more than 0.5 are relatively rare: Only 84 observations meet this criterion. Table 1 provides an overview of these cases.

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-populistchanges
#| tbl-cap: Constitutional changes under populist governments (weighted populism score > 0.5)

# Read the .tex file content
tex_content <- readLines("results/tables/populistchanges.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

<!--# In Europe, with the exception of Italy and Malta, constitutional changes by populist governments seem to be a rather Eastern European phenomenon. Also notable is that it is mainly right-wing populist governments that have been able to implement constitutional changes. A different picture emerges in Latin America. Here, constitutional changes have been implemented predominantly by left-wing populist governments. Since not all governments can be clearly classified on a left-right scale, we refrain from modeling the effect of left- and right-wing populism on the quality of democracy separately. -->

Figure 1 shows the distributions of the populism index in Europe and Latin America. Populist governments are not exceptional in Latin America). In Europe, the distribution is clearly skewed to the right, with a majority of governments with few populists. The majority of European country-year observations have a relatively low populism score (less than 0.5).

![Distribution of the Populism Score of Governments in Europe and Latin America](results/graphs/histogram_populistgov.pdf){fig-align="center"}

To analyze whether constitutional changes by populist governments have an effect on democratic quality, we use a country fixed-effects panel model. We use the dependent variables one year ahead ("lead") for each of the democratic dimensions.[^1] A constitutional change in one year will probably not have an impact on the democracy index in the same year, but only with a slight delay.

[^1]: Models for different numbers of leads can be found in @tbl-leadlibdem to @tbl-leadegaldem .

# 4. Results

The results shown in @tbl-results confirm our hypothesis that constitutional changes by left- and right-wing populist governments have different implications for democratic quality.[^2] According to the data changes by left-wing populists have had a positive impact on liberal, electoral and egalitarian democracy. For right-wing populists, we do not see any significant impact on democratic quality at all. @fig-interaction shows the average marginal effect of a constitutional change for conditioned by populism and government ideology. The figure only shows the effect for democratic dimensions where changes under populist governments have had a positive effect, the figures for civil society and participation can be found in the appendix (@fig-interactionapp).

[^2]: The results are robust for different leads (1-4), see appendix @tbl-leadlibdem to @tbl-egaldem.

The results imply that, so far, our understanding of constitutional changes under populist governments are too informed by few salient cases such as Hungary. Once we look at the bigger picture, we do not find the expected negative effect on liberal democratic institutions and polyarchy.

Considering the ideology of left-wing populists it does not come as a surprise that egalitarian democratic ideals are improved under these governments. Again, we do not find the same effect for right-wing populist governments.

\blandscape

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-results
#| tbl-cap: Main Regression Models

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_main.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

\elandscape

With its emphasis on popular sovereignty and the implementation of the people's will, one would expect that populists in office implement more possibilities for popular participation. Our results show that neither left- nor right-wing populists have improved the quality of participation or civil society through constitutional changes.

![Average marginal effect of constitutional change conditioned by government ideology and government populism score.](results/graphs/change_effect.pdf){#fig-interaction fig-align="center"}

Of course, left-wing populists in power in Europe have been rare to date. The positive effect of left-wing populists could be driven by the relatively young nature of Latin American democracies. If liberal democratic institutions are not only developing, populist parties in power might have a more positive impact on these institutions [@RuthLovell.2022]. We test this by replacing the ideology dummy with the democratic quality ahead of the respective country-year observation (lag of 2). @tbl-dynamic in the appendix shows that the triple-interaction effect is not significant in any of the models.

## 4.3 The Content of Constitutional Amendments by Populist Governments

So far, our analysis has focused on changes in the liberal democracy index or the civil society index. To gain a more detailed impression of the content of constitutional amendments by populist governments, we have taken a further step by categorizing the content of constitutional amendments. The strength of the executive is measured by an index developed by @Melton.2014. This measures whether the executive has the power to initiate legislation or constitutional amendments, issue decrees, declare a state of emergency, as well as enforce its power over other institutions through veto power, and have rights reviewed for constitutionality or dissolve parliament. The index of independent judiciary rights is based on @Melton.2014 and measures the number of constitutional norms that strengthen an independent judiciary.[^3] The index of political rights includes the guarantee of freedom of expression, as well as freedom of assembly, science, press, strike and trade union rights. Social rights include the guarantee of a certain standard of living, health protection at work, financial support, social security, and the right to a fair trial.

[^3]: Included are the independence of the judiciary in the constitution, whether at least two actors are involved in the nomination and appointment of judges to the Constitutional Court, whether the dismissal of judges is severely restricted and limited only to serious misconduct or constitutional violations, and whether judges' salaries are protected. Instead of including lifetime appointments, we include whether the re-election of judges is excluded.

![Content changes of constitutional amendments by populist governments (government weighted populism score \> 0.5) that led to democratic regression.](results/graphs/rights_change.pdf)

If we now look at the content of populist (weighted populism score \> 0.5) constitutional amendments that have led to a deterioration in democratic quality (change in the V-Dem index on liberal democracy compared with the previous year \< 0), these categories reflect the complexity of the relationship between populism and democratic quality discussed earlier. In seven cases, populist governments made constitutional changes that we can capture with the indices described and that led to a weakening of liberal democracy. But these examples already give us a sense of the relationship between populism and constitutional regression. While the power of the executive is always strengthened (executive aggrandizement), the picture is different when judicial independence is curtailed. In some cases, it is curtailed, but in others judges are granted more rights. Often, the social component of the constitution is also improved. This Janus-faced nature of constitutional amendments by populists is repeatedly emphasized in the literature and is also evident when we analyze the constitutional content itself [@Mudde.2013b; @Mudde.2012b; @Blokker.2019d; @LaTorre.2020].

However, the cases also show the importance of reasoned case selection. Hungary, which is often used as an example, differs from other cases of populist parties in government because here not only is the executive strengthened, but social rights are also weakened. If we rely on one-sided or false examples without justification, we easily overlook the complexity of the relationship between populism and democratic quality.

## 4.4 Case Studies: Democratic Regression with and without Constitutional Amendments

Our Large-N investigation is limited to constitutional regression. Accordingly, the question arises whether democratic regression is always accompanied by constitutional regression. We cannot answer this question in its entirety in this article. However, initial conclusions can be drawn by comparing Poland and Hungary.

Both countries have been the focus of discussion of democratic regression under populist governments in recent years. In Poland, the national conservative Prawo i Sprawiedliwość (PiS) has ruled in various coalitions since 2015. In Hungary, the equally national-conservative Fidesz-Magyar Polgári Szövetség (Fidesz), headed by Victor Orbán, has been in government since 2010. In many respects, developments in Poland and Hungary have been similar in recent years. In both countries, the governments have set themselves the goal of establishing an "illiberal" form of democracy. The Polish PiS even sees Hungary as a model for its own restructuring of Polish democracy."[@Sadurski.2019]

The similar orientation of Fidesz and PiS can also be seen in the data. Figure 5 shows that since the changes of government in Hungary in 2010 and Poland in 2015 -- each represented by a vertical dashed line -- the countries are classified as much more populist than before. In Hungary, the populism index jumps from 0.31 to 0.75 in 2010, and in Poland from 0.22 to 0.98 in 2015, near the maximum of the scale. Poland already had a populist government led by the PiS in 2005 and 2006.

The similar orientation of Fidesz and PiS can also be seen in the data. Figure 5 shows that since the changes of government in Hungary in 2010 and Poland in 2015 -- each represented by a vertical dashed line -- the countries are classified as much more populist than before. In Hungary, the populism index jumps from 0.31 to 0.75 in 2010, and in Poland from 0.22 to 0.98 in 2015, near the maximum of the scale. Poland already had a populist government led by the PiS in 2005 and 2006.

The similar development of the Democracy Index seems to indicate a similar approach by both populist governments to the restructuring of democratic institutions. Figures 5 and 6 show the years with constitutional changes as gray vertical lines next to the course of the Democracy Index. While the decline in the Democracy Index in Hungary clearly corresponds to constitutional changes by the Fidesz government, this is not the case in Poland. One obvious reason for this difference is that PiS does not have the necessary two-thirds majority in the Polish parliament, the Sejm, to push through constitutional amendments. Orban's Fidesz in Hungary, on the other hand, won an absolute majority of votes in 2010 and, due to the electoral system, a two-thirds majority of seats in parliament, which it confirmed in subsequent elections. Equipped with the necessary majority, a new constitution was adopted in 2011, which was amended several times in the following years.

Although PiS was denied such a transformation of the Polish constitution due to the lack of a parliamentary majority, the regime there has found a different strategy to achieve similar effects. Central to Poland's shift away from liberal democracy is the occupation and reshaping of the Constitutional Tribunal, which oversees constitutional compliance in Poland. In 2015, President Duda refused to administer the oath of office to three constitutional judges who had been confirmed by parliament prior to the October 2015 parliamentary elections. The vacant posts were then replaced after the elections by judges nominated by the PiS government. In other changes to the Constitutional Tribunal, the post of president was filled, among others, by judge Julia Przylebska, who is close to PiS. The office includes far-reaching powers to appoint panels of judges and to designate the respective rapporteur. A critical counterpoint to this is that the reorganization of the Constitutional Tribunal allowed the PiS government to pass a number of laws that objectively violated the constitution but were not named as violations by the Constitutional Tribunal, which was paralyzed by strategic appointments."[@Sadurski.2019].

It is remarkable that there is an almost identical decline in the democracy index in both countries, behind which, however, lie different strategies of the governments of the two countries. If the same effects can be achieved by simple legislative changes, the question arises why the distinction between constitutional and simple legislative changes is so important. We argue that the distinction is particularly important because of the sustainability of the legislative changes. There is a qualitative difference between a constitutional amendment (Hungary), which can only be reversed with a qualified majority, and a strategy based on simple legislation (Poland), which could also be reversed with a simple parliamentary majority after a change of government. This qualitative difference is difficult to represent in the Democracy Index. Likewise, the analysis shows that constitutional changes are not a necessary condition for a significant decline in the quality of democracy. Democratic and constitutional regression thus do not always go hand in hand.

# 5. Conclusion

In this article, we have examined the extent to which democratic regression can be attributed to constitutional change by populist governments. We introduced the concept of constitutional regression for the strategic use of constitutional change to degrade democratic quality. In our approach, this can have an impact on both participatory and liberal democratic elements of a democracy.

Our results show that there are no clear correlations between constitutional regression and populist governments. Thus, our results contrast with the thesis that the presence of populist parties in governments always has a negative effect on the quality of democracy. While in Europe constitutional changes by populist governments have a negative effect on the quality of democracy, a positive effect can even be observed in Latin America. This contrasts with case studies of countries, for example, Hungary, ruled by Orbán, in which a fundamental restructuring of democratic institutions has taken place as a result of constitutional changes by the ruling Fidesz.

Our findings also confirm the warnings against drawing too quick conclusions about the relationship between populism and constitutional change. Various authors repeatedly point to the heterogeneity of populist parties and to the fact that populism can also have positive effects, for example on participation [@Blokker.2019c; @LaTorre.2020; @Mudde.2013b]. Our analysis also shows that conclusions about a clear relationship between populism, constitutional change and democratic regression are also not possible due to the rarity of constitutional reforms. In our Large-N study, it is clear that the Hungarian experience is not transferable to populist parties in Latin America.

Despite these limitations, we also find clear patterns. As already pointed out by Bermeo, the core of democratic regression lies in executive aggrandizement [@Bermeo.2016]. At this point, constitutional changes by populist governments that lead to democratic regression are also similar. These reforms also partially strengthen social rights; however, what the amendments very often overlap in is that power is centralized in the executive branch.

In the interpretation of our results, it must be noted that constitutional amendments are relatively rare events and are usually implemented only with a qualified majority in parliament, which populist parties often do not possess. Therefore, when desired constitutional changes are sought, opposition parties must also be involved. The comparison between Poland and Hungary has shown that populist governments can bring about significant changes even without constitutional amendments through reforms, reshuffling of the constitutional court, and legislation below the constitutional level.

### References

::: {#refs}
:::

 {{< include appendix.qmd >}} 
