---
title: 'Who Shields the Guardian: Does Trust in the Judiciary Constrain Court Curbing?'
subtitle: ""
author:
  - name: Jasmin Sarah König^[University of Hamburg. Jasmin was funded by a DFG Research Training Group – GRK 2503. Address for Correspondence jasmin.sarah.koenig@uni-hamburg.de]
    email: jasmin.sarah.koenig@uni-hamburg.de
    affiliation: University of Hamburg
  - name: Sebastian Block^[Leibniz University Hannover]
date: last-modified
date-format: medium
geometry:
  - top=30mm
  - left=30mm
  - bottom=30mm 
  - right=30mm
  - heightrounded
linestretch: 1.5
tbl-cap-location: top
format:
  wordcount-pdf:
    fontsize: 12pt
    include-in-header:
      text: |
        \usepackage{setspace}
        \onehalfspacing
        \renewcommand{\abstractname}{Abstract}
        \usepackage{xcolor}
        \usepackage{booktabs}
        \usepackage{amsmath}
        \RequirePackage{xcolor} % Required for specifying custom colours
        \colorlet{mdtRed}{red!50!black}
        \definecolor{lightdark}{RGB}{224,227,223}
        \definecolor{lightlightdark}{RGB}{248,249,248}
        \definecolor{chaptercolor}{RGB}{36, 87, 71}
        \definecolor{slategray}{RGB}{119, 136, 153}
        \definecolor{darkslategray}{RGB}{89, 106, 123}
        \definecolor{colordark}{RGB}{125, 140, 120}
        \definecolor{colorcolorful}{RGB}{208, 157, 106}
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
editor: visual
bibliography: citation/lit.bib
csl: citation/ejpr.csl
classoption: abstract
---

```{r}
#| include: false  
#| 
library(xtable) 
library(papeR) 
library(stargazer) 
library(gt) 
library(modelsummary) 
library(tidyverse)
```

Courts are often described as the guardians of democracy [@Vanberg.2005; @Prendergast.2019; see @Boese.2021]. However, with the rise of populism, these guardians have become targets in need of protection. Once in power, populists frequently dismantle the independent judiciary [@Scheppele.2019; @Vosskuhle.2023; @Muller.2016]. The prevailing notion is that the fear of a public backlash prevents governments from undermining judicial independence [@Vanberg.2001; @Staton.2010; @Krehbiel.2020]. However, we lack empirical evidence across countries and over time to support this assertion, as well as the claim that populists in office attack judicial independence. This research note examines whether populists in power attack the judiciary more frequently across 40 countries and over a span of 20 years, and whether high public trust in the judiciary reduces the likelihood of such attacks.

Populism is characterized by a belief in popular sovereignty without constraints, viewing any actor that obstructs the people's right to power as an enemy [@Mudde.2004; @Abts.2007; @Mudde.2012b; @Mazzoleni.2020]. If the guardians of democracy are now in the crossfire of populist governments, the question arises: who can shield the guardian against attacks? Such attacks can, for example, take the form of replacing judges (court purging) or adding friendly judges (court packing) [@Arato.2019; @Huq.2018]. According to conventional wisdom, an independent judiciary should be protected by public support. The common notion is that citizens who value the judiciary as an independent arbiter will defend it in elections if it is under political threat [@Vanberg.2001; @Staton.2010; @Krehbiel.2020]. If this is true, the likelihood of court curbing should decrease with higher public trust in the judiciary.

Anecdotal evidence shows that once populists come into power, they translate their beliefs into attacks on the judiciary. In Hungary, Orbán undermined judicial independence by replacing judges and changing the Constitutional Court's jurisdiction [@AydinCakir.2023]. In Ecuador, the populist administration under President Moreno pressured judges to make decisions in their favor [@hrw.2018]. In Poland, the PiS government forced the retirement of judges who were not co-partisans [@Sadurski.2019]. To the best of our knowledge, only @Bartels.2023 has analyzed whether public support for democracy protects the judiciary from backlash. Their Kenyan case study found no effect for this mechanism. However, research on both the attacks on the judiciary by populists and the potential protection provided by public support has predominantly focused on single-country studies [see @AydinCakir.2023; @Mudde.2021; @granata.2022; @Magalhaes.2023; @Bartels.2023]. To generalize these findings, we need cross-country comparisons on the impact of populists in office on the judiciary.

We use a dataset including 40 European and Latin American countries to test these arguments. We constructed the dataset combining data from V-Dem [@vdemdata.2022], V-Party [@VParty.2022], Euro- and Latinobarometer [@EuropeanCommission.2020; @Latinobarometro.2023b] and a paper by @RuthLovell.2022. Our results show that the danger of attacks on the judiciary, court packing and court purges rises significantly if populists are in power. However, we do not find any evidence that an increase in trust in the judiciary decreases the likelihood of court purges or packing, neither under populist nor non-populist governments.

## The Rise of Populism and Why it Threatens the Independent Judiciary

Data shows independent judiciaries increasingly find themselves under threat in Europe and North America [@vdemdata.2022, appendix @fig-attacks]. This trend is frequently attributed to the rise of populism [@Arato.2019; @Scheppele.2019; @Muller.2016; @Schafer.2021]. We frequently observe that populists around the world undermine judicial power once they are in government. The populist Kirchner government in Argentina tried to push a senior judge out of the court [@Scribner.2017]. Trump attacked federal judges while in office and established a Republican majority in the Supreme Court [@iba.2024; @Nemacheck.2022]. Additionally, a similar picture arises for Hungary [@AydinCakir.2023], Poland [@Sadurski.2019], and Ecuador [@hrw.2018] as described in the introduction.

However, we do not observe such a pattern in every case of governments with populist participation. While SYRIZA was in government between 2015 and 2019, judicial independence did not significantly decrease. Indeed, their constitutional reform proposal even included the implementation of a Supreme Court to strengthen the rule of law [@Lorencka.2019].

So far, most cross-country studies have investigated the effect of populists in power on democratic quality more generally, with judicial independence as a part of an index [@Huber.2017; @RuthLovell.2022; @Juon.2020]. Their results paint a similar picture: Populists in power decrease the quality of liberal democracy.

The predominant findings that populists in power have a negative effect on democratic quality is in line with the ideational approach to populism. According to the widely used approach, populism is defined as a thin-centered ideology [@Mudde.2004]. The populist ideology builds on the belief in a homogeneous people that is morally good and capable to govern itself. In contrast, elites are depicted as the enemy of the morally good people [@Mudde.2004]. The populist vision of a democracy governed by majority rule directly challenges the system of checks and balances integral to liberal democracies [@Canovan.1999; @Meny.2002b; @Abts.2007].

The judiciary is tasked with safeguarding citizens from governmental abuses of power, as well as shielding minorities from potential tyranny by the majority [@Rogers.2020; @Stephenson.2004; @StoneSweet.2002; @Whittington.2010]. By independently interpreting and applying the law, the judiciary acts as a check on the executive and legislative branches. Thus, the judiciary, in particular constitutional courts, are often regarded as the guardians of democracy [@Ferejohn.2009; @Prendergast.2019; @Vanberg.2005].

Given the judiciary's pivotal role as a check on the executive, the weakening of judicial power is often one of the first steps in the process of democratic backsliding [@Satrio.2018; @Ginsburg.2018b; @Khaitan.2019]. A judiciary that is either captured or weakened has less options, or inclination, to stop subsequent attempts to erode democracy. Goverments have a whole range of options to target the independent judiciary. They can attack either the court's composition, or its jurisdiction. We focus on the former case: By replacing (purging) or adding friendly judges (packing) governments can capture a court [@Arato.2019; @Huq.2018]. We expect that this is more frequently the case if populists are in power.

If a government lacks the majority to undermine judicial independence, it can still use words as weapons to set the stage for future court-curbing measures. Courts rely heavily on their perceived legitimacy among the public, making them vulnerable to political attacks that portray them as biased, corrupt, or incompetent [@AydinCakir.2023; @vdemcodebook.2022].

\begin{quote} \textbf{Hypothesis 1}: Populist in power increase the likelihood that court packing or court purging takes place. \end{quote}  \begin{quote} \textbf{Hypothesis 2}: Populist in power increase the likelihood that verbal attacks on the judiciary take place. \end{quote}

## Who guards the guardian?

If populists attack the guardians of democracy, one must wonder whether anyone or anything shields the guardian from such threats. Scholars are increasingly interested in what makes democracies resilient against backsliding [@Merkel.2021; @Boese.2021; @Laebens.2021; @LaTorre.2018]. Most studies agree that an organized and active resistance against autocratization among citizens is crucial to avert the erosion of democratic institutions, such as the judiciary [@Druckman.2023; @Laebens.2021; @Luhrmann.2021].

This notion is also prevalent in the literature on judicial independence. Citizens are often regarded as the shield of the independent judiciary [see @Vanberg.2015]. While being a crucial democratic institution, the judiciary is also inherently weak. To unfold its power, the judiciary needs support among the public [@Buhlmann.2011; @Vanberg.2001; @Staton.2006]. Particularly high courts have neither the power of the "purse" nor the "sword". Still, they have to take decisions against the preferences of the majority and the executive [@Gibson.1998]. One common explanation as to why governments respect judicial independence, despite the possibility to simply ignore the judiciary, is the fear of a public backlash [@Vanberg.2001; @Vanberg.2005; @Krehbiel.2020]. This idea has been developed primarily in the realm of judicial review. The notion is that if citizens support and trust the judiciary, they will defend the institution in upcoming elections by voting for a different party [@Vanberg.2001; @Vanberg.2005; @Gibson.1998; @Staton.2006; @Staton.2010; @Krehbiel.2020].

The notion that public support shields the judiciary has been under empirical scrutiny. Recent research examines the willingness of citizens to translate high support for democratic norms into action at the ballot box if incumbents behave undemocratic. The evidence from these studies is mixed. Some studies imply that voters are not willing to punish attacks on judicial independence if the partisan or policy costs are too high [@Driscoll.2023b; @Jacob.2024; @Wunsch.2023; @Bartels.2020; @GRAHAM.2020; @SVOLIK.2020; @Lewandowsky.2023]. Instead support for (horizontal) judicial power depends on whether their preferred party is in power [@Bartels.2020c]. If the incumbent is a co-partisan, support for horizontal judicial power decreases. Likewise, @Bartels.2023 find that support for the judiciary depends on whether judicial decisions are made in one's own interest. But, the evidence is not unilateral. Other studies find that candidates who attack the independent judiciary receive a lower vote share [@Engst.2023]. @Saikkonen.2023 find that this is the case, even if the candidate represents the respondent's preferred policy positions on a contentious topic. @Krehbiel.2020 further shows that voters did punish non-compliance with judicial review in a cross-country study of 74 states. Still, we lack data beyond Western democracies that compares the effect of citizens attitudes on court-curbing.

If public trust in the judiciary acts as a shield, then we should observe fewer attacks on the judiciary as public trust increases. Governments concerned about potential backlash would view each citizen who trusts the judiciary as a potential adversary likely to punish the government for undermining judicial independence. Consequently, we expect that attacks on the judiciary, including court-packing, purging, and verbal attacks, will occur less frequently when trust in the judiciary is high.

\begin{quote} \textbf{Hypothesis 3}: With increasing trust in the judiciary, the likelihood of court packing or court purging decreases. \end{quote}  \begin{quote} \textbf{Hypothesis 4}: With increasing trust in the judiciary, the likelihood of verbal attacks on the judiciary decreases. \end{quote}

<!--#Combining the two strands of research, the question arises whether public trust in the judiciary can act as a shield against populist attempts to undermine judicial independence. So far, we know little how populist voters view the judiciary. While some studies show that populists citizens hold similar democratic attitudes as non-populist citizens [@Wuttke.2022; @Zaslove.2023]. According to @Wuttke.2022, populist citizens support the rule of law similarly as other citizens. However, such studies do not account for the incumbent effect. Once populists are in office, the support for non-majoritarian institutions might decrease among populist citizens [@Muller.2016; @vanderBrug.2021]. We test whether the likelihood that populists attack the judiciary is conditioned by trust in the institution.   -->

<!-- #Citizens particularly important in young democracies where institutions are still volatile [@]. These are likely also the countries where trust is low. -->

<!-- @Bermeo.2016 (p. 10-11) summarizes the core of democratic backsliding as executive aggrandizement, in the course of which horizontal accountability is weakened and power is accumulated in the executive. This often includes the dismantling of judicial independence. -->

<!-- Strong horizontal accountability can increase democratic resilience as well [@Boese.2021, XXX]. Particularly the judiciary is often regarded as the "guardian" or "bulwark" of democracy [@Ferejohn.2009; @Satrio.2018; @Boese.2021]. Courts, particularly those that can exercise judicial review, hold the power to restrain the executive [@Ginsburg.2018b]. But, institutions that can hold the executive accountable, such as an independent judiciary, are frequently restricted in the process of democratic backsliding themselves. The power of an independent judiciary places it in the crossfire of autocratization tendencies. -->

<!-- Under this crossfire, the judiciary can only prevail if it is supported by the public  -->

## Research Design

### Data & Models

To test our hypotheses, we created a new dataset based on data from V-Dem [@vdemdata.2022], V-Party [@VParty.2022], Euro- and Latinobarometer [@EuropeanCommission.2020; @Latinobarometro.2023b], and a paper by @RuthLovell.2022. The data entails observations from 40 countries in Europe and Latin America between 1998 and 2020.[^1] Overall, our analysis includes 746 country-year observations. Populist were in power in 112 of these cases.

[^1]: Because of missing data, the exact time-frame varies between countries (see appendix @tbl-cases).

To test our hypotheses, we employ multiple country fixed-effects models to account for country-specific differences. Additionally, we address heterogeneity in the time-series data by using heteroskedasticity-consistent standard errors [@Hayes.2007]. Our main analysis includes two models, one on court packing and purges, the other on verbal attacks on the judiciary.[^2]

[^2]: We run multiple robustness checks. To ensure the validity of our findings despite the possible heterogeneity in the time-series data we also run random-intercept models (see appendix, @tbl-random). We also include a robustness check with different imputations/no imputation of the trust variable (see operationalization, see appendix from @tbl-indepv_w_pp). We further test some alternative explanations: If populists want to undermine judicial independence, higher trust might only have an effect under non-populist governments. We test this with interaction effects (appendix, @tbl-results-appendix). Whether attacks on judicial independence occur more often once such attacks have started (once the dam has broken, a government might not fear a backlash anymore - see appendix @tbl-results_appendix). Lastly, we test whether the level of judicial independence has an effect on the likelihood of attacks on the judiciary (see appendix from @tbl-dynamic_ji_pp).

### Operationalization

To test our hypotheses we use two different dependent variables. For hypotheses I and III our dependent variable indicates whether a country has experienced politically motivated changes in court compositions. The V-Dem dataset collects this information building on aggregated expert jugdements. The two variables of interest are on judicial purges[^3] and court packing[^4] which indicate whether there was a "massive" or "limited, politically motivated increase in the number of judgeships" [@vdemcodebook.2022, p. 163], or "removal of judges" [@vdemcodebook.2022, p. 164].[^5] The expert assessment is converted from an ordinal to an interval scale, thus the variables are continuuous on a scale from -5 to 5 [@vdemmethod.2024]. We combine the variables on packing and purges into one variable `Court Purges & Packing`, by choosing the score that indicates more political interventions (see appendix @fig-densitydv). The higher this value, the more often court purging and packing took place according to the experts.

[^3]: `v2jupurg_ord` in the original dataset.

[^4]: `v2jupack_ord` in the original dataset.

[^5]: See the appendix for robustness tests with different operationalizations of the variable.

For hyothesis II and IV, we use V-Dem's variable on government `Attacks on the Judiciary`. The variable indicates whether governments frequently verbally attack the judiciary, for example through "claims that it is corrupt, incompetent or that decisions were politically motivated" [@vdemcodebook2022, p. 164].[^6]

[^6]: `v2jupoatck` in the original dataset.

Our first explanatory variable whether a government was `Populist` builds on data from @RuthLovell.2022 who combined the PopuList [@Rooduijn.2019], data from @Huber.2017, and their own coding. We have expanded the dataset for the years 2019 and 2020. The variable is binary with 1 indicating that the head of government is populist, otherwise the variable is 0.

The second explanatory variable on `Trust` in the judiciary is aggregated from the Latinobarometer [@Latinobarometro.2023b] and the Eurobarometer [@EuropeanCommission.2020]. From the survey data, we calculated a variable that indicates what percentage of respondents in the respective year trusted the judiciary in their country.[^7] Both, Latino- and Eurobarometer, include some years with missing observations. In these cases we have imputed data. Our imputation mechanism assumes that trust has developed linearly throughout the missing observations.[^8] In the main model, we use the last three years' mean for this variable (Trust~t-1,-2,-3~).[^9]

[^7]: While the Eurobarometer only includes two possible answers, *trust* or *do not trust*, the Latinobarometer includes four categories "A lot", "Some", "Little" and "No trust" [@Latinobarometro.2023b]. To ensure comparability between the Euro- and Latinobarometer, we have recoded the variable into a binary variable that codes people who trust the judiciary *a lot* or *some* as trusting the judiciary. We run a robustness check in which only respondents who trust the judiciary *a lot* are included in the share of people who trust the court. Since the regression models include country fixed-effects, differences in the coding between the continents should not lead to wrong estimates as only the within country variance of the variable is used.

[^8]: See the appendic for robustness checks with other operationalizations of the trust variable (see @tbl-indepv_w_atck and @tbl-indepv_w_pp).

[^9]: See the appendix for robustness checks with different lags (see @tbl-indepv_w_pp and @tbl-indepv_w_atck).

We control for several factors: Whether governments undermine the judiciary could be affected by their legal possibilities to do so. We include an `Executive Power` index build with the Comparative Constitutions dataset that measures the power of the executive within a country [@ComparativeConstitutions.2005; @Elkins.2012]. Since executive aggrandizement is more likely to occur in a presidential system [@Voigt.2020], we include a dummy on `Presidential Systems`in our models. In a similar vein, we control for the share of `Surplus Seats` a government has since it is easier to find the necessary majority for such drastic attacks on the judiciary if a government has more seats. Governing in a `Coalition`can moderate the possibility to undermine the judiciary, since multiple parties need to agree. Both variables were built based on data from the V-Party dataset [@VParty.2020]. Further, we include the average `GDP` growth from the last three years based on World Bank data [@wdi.2020].

## Results

In this section, we present the results of our analyses. First, we show descriptively how often attacks on the judiciary occur and test whether attacks are more frequent under populist rule than under non-populist governments. We then use multivariate analysis to test all four hypotheses of this research note.

*Descriptive analysis*

@fig-boxplot shows the frequency of court curbing and packing as well as oral attacks per year on the judiciary, comparing populist and non-populist governments. The results show that populist governments tend to use court curbs more often than non-populist governments. On average, court curbing is more frequent under populist rule (0.15 on the V-Dem interval-point scale) than under non-populist governments (-0.91 on the V-Dem interval-point scale). This difference is significant (p\<0.01) in a one-tailed t-test. These results provide descriptive support for the first hypothesis. The graphic also suggests that populist governments tend to attack and replace the judiciary more often than non-populist governments. On average, oral attacks on the judiciary are less likely under non-populist rule (0.63 on the V-Dem interval-point scale) than under populist rule (0.84on the V-Dem interval-point scale ). Again, this difference is significant (p\<0.01) in a one-sided t-test. The descriptive results also are in line the second hypothesis since they show that populist parties tend to orally attack the judiciary more often compared to non-populist governments.

![Expert assesment of court curbing & packing or verbal attacks on the judiciary under populist and non-populist rule. Interval-point scale from @vdemdata.2022, see also @vdemmethod.2024. Whiskers represent 1.5\*IQR.](results/graphs/populism_attacks.pdf){#fig-boxplot fig-align="center"}

*Regression analysis*

The findings from our multivariate regression models corroborate the initial hypothesis that instances of court packing and judicial purges are more prevalent under populist governments (see @tbl-results, model 1). The same is the case with regard to attacks on the judiciary (see @tbl-results, model 2). Both models are run using robust standard errors. For both dependent variables, the coefficient associated with populism is positive and significant across all models. These results reinforce the prevailing notion that populists in government undermine judicial independence and are thus in line with our first and second hypotheses.

\setlength{\tabcolsep}{3pt}

```{r}
#| output: asis  
#| echo: false  
#| warning: false  
#| label: tbl-results  
#| tbl-cap: OLS Regression Models    

# Read the .tex file content  
tex_content <- readLines("results/tables/mainmodels_small.tex")  %>%    gsub("Trustmean3", "Trust$_{\\bar{t}_{-1,-2,-3}}$", ., fixed = TRUE) %>%   gsub("0.001", "0.001", ., fixed = TRUE)  

# Print the content as raw LaTeX  
cat(tex_content, sep="\n")
```

However, our results raise doubts whether the public's attitudes toward the court act as a shield for the judiciary. The analysis indicates that an increase in citizens who trust the judiciary does not reduce the likelihood of court packing and purges, nor of attacks on the judiciary as stated in hypotheses 3 and 4. In none of our models do we find evidence that an increase of trust within a country decreases the likelihood of court purges or packing. Our model shows a significant positive effect of higher trust what implies that court purges and packing become even more likely when trust is high (see @tbl-results, model 1). However, the significance of the coefficient is not robust across different operationalizations of the variable (see appendix @tbl-indepv_w_pp). In all of the models we calculated, trust does not decrease the likelihood of court purges and packing. Overall, our results indicate no reliable effect of trust on court packing and purging. In addition, we find no signficant effect of trust affecting attacks on the judiciary. All our models show that higher trust among the public does not decrease attacks on the judiciary (@tbl-results, model 3, see appendix @fig-coefpp and @fig-coefatck).

## Discussion & Conclusion

According to the literature populists endanger courts [@Muller.2016; @Scheppele.2019; @Vosskuhle.2023; @Arato.2019] which are often perceived as the guardians of democracy [@Prendergast.2019; @Vanberg.2005; @Boese.2021; @Albertazzi.2013]. However, research so far did not analyze this phenomenon on a large scale across multiple countries and for a long time period. In this article, we show that populists in government are indeed a danger to judicial independence, as depicted in many recent writings [@Arato.2019; @Scheppele.2019; @Muller.2016]. Our findings corroborate for a large dataset, including 40 countries across multiple years, what scholars previously found in in-depth single country case studies: Populists attack the judiciary verbally and undermine its independence by changing court compositions [@LaTorre.2020; @Mazzoleni.2020; @Scribner.2017; @Rogenhofer.2020; @LaTorre.2018; @AydinCakir.2023; for case studies see @Hoffmann.2022]. Furthermore, we tested whether the common assumption that high public support could be a shield for the guardian that preventing packing, purging and attacks of courts [@Vanberg.2001; @Staton.2006; @Krehbiel.2020]. However, we do not find that higher public trust in the judiciary has a negative effect on either court packing and purging or attacks on the courts. Thus, we find no empirical support for the common assumption in the literature that public trust can protect courts.

Our findings contribute to the ongoing debate on democratic backsliding with a focus on the erosion of judicial independence. By looking at 40 states across Europe and Latin America, we show that high trust in the judiciary alone does not create a shield that protects the institution. Our results show that citizens across Europe and Latin America do not act as a shield for the judiciary. Our findings are in line with recent evidence that voters lack the willingness to translate their diffuse support for the court into actions at the ballot box [@Jacob.2024; @Wunsch.2023; @Driscoll.2023b]. Similary, studies have shown that voters are willing to trade of judicial independence for other factors such as economic growth, partisanship or policy congruence [@Driscoll.2023b; @GRAHAM.2020; @Lewandowsky.2023]. These studies might explain why governments still undermine judicial independence despite increasing trust in the institution among the public.

However, there might also be other reasons for our finding. Short-term increases in trust might not have the same effect as long-term trust in the judiciary. A court that has been a trusted institution for multiple decades might be defended more willingly than a comparatively young institution, or one that has only gained the trust of the public. Another reason for this finding might be that the public is uncertain whether an attack on the judiciary has or will occur [@Chiopris.2024; @Jacob.2024]. A study by @Magalhaes.2023 implies that voters in Hungary did not perceive the judicial reforms under Orbán as problematic. Their results show that trust in the judiciary in Hungary did not decrease once the judiciary's independence was undermined. @Jacob.2024, on the other hand, finds that Polish voters clearly recognize undemocratic behavior but are not willing to withdraw their votes at the ballot box.

So far, experimental studies are restricted to few countries. Our cross-country comparison corroborates the findings of earlier research that the effect of voters' support as a shield for the judiciary might have been overestimated. The results show that we need more research on what drives citizens to punish incumbents' undemocratic actions beyond partisanship and policy preferences, and whether there is any way to increase citizens' willingness to defend democratic institutions.

{{< pagebreak >}}

### References

::: {#refs}
:::

{{< pagebreak >}}

{{< include _appendix_koenig_block.qmd >}}
