---
title: "CONSTITUTIONAL CHANGES UNDER POPULIST GOVERNMENTS"
subtitle: "An Empirical Analysis"
format: pdf
editor: visual
documentclass: article
author:
  - name: Jasmin Sarah König^[University of Hamburg. Jasmin wurde durch die Deutsche Forschungsgemeinschaft (DFG) – GRK 2503 gefördert. Address for Correspondence jasmin.sarah.koenig@uni-hamburg.de]
    email: jasmin.sarah.koenig@uni-hamburg.de
    affiliation: University of Hamburg
  - name: Tilko Swalve^[Leibniz University Hannover]
bibliography: citation/lit.bib
csl: citation/ejpr.csl
geometry:
  - top=30mm
  - left=30mm
  - bottom=30mm 
  - right=30mm
  - heightrounded
linestretch: 1.5
classoption: abstract
abstract: |
  Populism and liberal democracy are – at least in parts – in conflict. Researchers have discussed a possible relationship between populist parties in government and democratic regression. In countries with sweeping populist majorities - such as Hungary - populists in power have undermined democratic institutions through constitutional changes. Our paper analyzes whether this mechanism is systematically applied by populists in government in Europe and Latin America. Using V-Dem and V-Party data, we investigate whether constitutional change leads to democratic regression more often under populist governments. The results from our multi-level model show that the relationship between populism and constitutional regression is ambiguous. 
include-in-header: 
  text:
    \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    \renewcommand{\abstractname}{Abstract}
    \usepackage{booktabs}
    \usepackage{siunitx}
    \newcolumntype{d}{S[
      input-open-uncertainty=,
      input-close-uncertainty=,
      parse-numbers = false,
      table-align-text-pre=false,
      table-align-text-post=false
      ]}
---

```{r}
#| include: false
options(tinytex.verbose = TRUE)

```

<br>

{{< pagebreak >}}

Put this somewhere?\
In the wider literature, however, the relationship between democracy and populism is much more controversial [@Canovan.1999; @Meny.2002b; @Mouffe.2005; @Laclau.2005; @Abts.2007; @Mudde.2012b; @KONIG.2022b]. While many authors agree that populism and liberal democracy are at least partially in conflict, they frequently also point out that populism may well have a corrective effect on a less responsive, highly formalized democracy [@LaTorre.2020; @Mudde.2012b; @Canovan.1999]. Both @Blokker.2019d and @Bugaric.2019 refer to the variance of populist parties and argue that populist constitutional amendments do not only have negative effects on the quality of democracy. Solely a preference for frequent constitutional amendments is not a threat to democracy [@Tushnet.2021, p. 73].

## 1. Introduction

In modern democracies, constitutions are implemented to safeguard democratic institutions, processes, and values [@Habermas.2001]. But, constitutions do not only constrain public authority and protect citizens from state power [@StoneSweet.2002]. It often also constrains popular sovereignty by introducing checks and balances, or minority rights. The populist ideology, on the other hand, revolves around popular sovereignty. Populists vow to give the power back to the people. We analyze whether populists use constitutional changes as a mechanism to adapt the constitution to the will of the people. In a second step, we estimate the effect of such changes on different levels of democratic quality.

The relationship between populism and constitutionalism has not been in the focus of research for long. Some scholars argue that populism aims to democratize constitutions by increasing participation and adapting the contents of the constitution to the popular will, we call this the ideological approach [@Blokker.2019c; @Tushnet.2021]. In populism, law is understood as the result of the "will of the people" and thus as political [@Blokker.2019]. They regard the constitution as a living document instead of a rather static safeguard of set democratic norms. Others argue that populism has an instrumental approach to constitutionalism. According to this approach, populists adapt the constitution in way that consolidates their power [@Mudde.2021; @CristobalRoviraKaltwasser.2013; @Muller.2017]. We argue that, if populists are indeed acting ideologically, we just observe an increase in constitutional changes under populist governments. But, if populists merely try to consolidate their power through constitutional changes, the constitution does not need to be changed more frequently. Few significant changes are sufficient to reach such a goal.

Based on the instrumental approach many scholars connect populists in power with democratic backsliding [@Schafer.2021; @Ginsburg.2018; @Diamond.2021]. More precisely, they argue that populists use constitutional changes to undermine democratic institutions [@Landau.2013; @Scheppele.2018; @Huq.2018; @Muller.2017b; @Muller.2017]. . The populist imperative to enforce the will of the majority are at odds with the preservation of minority rights and the influence of non-majoritarian institutions in liberal democracies [@Abts.2007]. However, democratic regression or progress happens not only with regard to liberal democratic institutions but also on other democratic dimensions, such as participation or egalitarianism.

We argue that the relationship between constitutional change under populist governments and democratic quality is not as clear as the literature on democratic regression suggests [@Landau.2013; @Scheppele.2018; @Huq.2018]. While populism is in conflict with liberal democratic institutions, it can have a positive impact on participation or egalitarianism [@Mudde.2012b]. Similar to a study by @RuthLovell.2022, we analyze the effects on democratic quality on distinct democratic dimensions, liberal democracy, polyarchy, egalitarianism, civil society and participation. We further test whether this effect differs between right-wing and left-wing populist parties in power.

Using data from @RuthLovell.2022, the V-Dem [@Vdem.2021] and V-Party [@VParty.2022] as well as the Comparative Constitutions Project datasets [@ComparativeConstitutions.2005], we examine first examine whether constitutional changes are more likely to occur under populist governments. Our results show that populists in office do not change the constitution more frequently than other governments. We argue that this speaks against the understanding of a purely ideological relationship between populism and constitutionalism.

We then estimate what effects constitutional change under populist governments have had on democratic quality. The results imply that populists do not necessarily use constitutional changes to aggrendize power. Instead, we find that the effect of constitutional amendments by populist parties in power differ between democratic dimensions and party ideology. Our results show that constitutional changes under left-wing populist governments have had a significant positive effect on liberal democratic institutions, polyarchy and egalitarianism. For right-wing populist parties, on the other hand, we only find a significant negative effect on polyarchy. These findings are in contrast to well-known cases, such as the constitutional changes in Hungary that have had significant effects on liberal democratic institutions. We argue that our results show that we should not infer from few well-known cases to all populist governments. Instead we need to consider the heterogeneous nature of populists in power.

## 2. POPULISM & CONSTITUTIONALISM

The most commonly used definition of populism in Political Science describes it as a thin ideology [@Mudde.2004]. Populism usually emerges tied to a "denser" host ideology, for example fascism or socialism, and yet has its own ideological core. @Abts.2007 (p. 409) describe this core as a question of the distribution of power. The populist ideology is based on the majority principle, in which only the "will of the people" is to be implemented. Any constrains on this process are understood as an obstruction of democracy [@Abts.2007; @Mazzoleni.2020]. According to the populist ideology, the will of the people is homogeneous and therefore clearly recognizable [@Abts.2007; @Mudde.2004].

In its endeavor to implement full popular sovereignty, populism divides society into an evil elite and the good people [@Mudde.2004]. According to populists, the people recognize what is best for them, an idea that is often mirrored in saying such as *common sense*. Whoever prevents this general will from being implemented is described as an enemy of the people.

Liberal democracy must always balance institutionalization and popular sovereignty [@Canovan.1999]. This inherent incongruence is always part of liberal democracies. One attempt to institutionalize and safeguard democracy is to enshrine certain rights, institutions and values into a constitution [@Habermas.2001]. In many liberal democracies separation of power as well as checks and balances are implemented in a constitution to prevent a centralization of power, or a tyranny of the majority [@XXX].

These safeguards are in contradiction to the populist ideal of democracy. Different to liberal democracy, populism has a clear alignment between the two poles of institutionalization and popular sovereignty. The "will of the people" must be implemented as quickly as possible and without obstacles. Institutions and norms serve the sole purpose of supporting this process, but must never hinder it \[\@\].

This is also mirrored in the populist understanding of the law. Populism is based on the supremacy of the political [@Mudde.2012, p. 17; @Blokker.2019d]. Law can accordingly only express the outcome of political processes, but can never justify their restriction. The claim of a neutral law that stands above the political process is not recognized by populism. Instead, law is seen as a purely political medium. The goal of directly implementing the alleged "will of the people" leads to a special relationship between populism and constitutionalism. The constitution is supposed to reflect the will of the majority of the people and is thus not seen as a firmly established institution that is rarely changed, but as a living document that is purely political [@Blokker.2020d; @Mazzoleni.2020].

Populists do not strive to abolish constitutions, but to re-politicize them in the sense of the alleged "will of the people."[@Mazzoleni.2020; @Muller.2017b]. In order to always reflect the will of the majority, constitutions should be easy and quick to change according to the populist ideal [@Fabbrizi.2020]. With this understanding of the constitution, the distinction between ordinary and constitutional law is also abolished within the populist ideology [@Blokker.2020d]. If constitutional law is no longer seen as a guideline in the everyday political process, but only as a form of expression of the political, it loses its elevated and particularly safeguarded position.

**The ideological approach**

However, scholars are divided about what this implies for constitutionalism once populists are in power. According to the first approach, populists act purely ideological once in office and aim to give the power back to the people [@Tushnet.2021; @Blokker.2019d]. This understanding of populism and constitutionalism is closely related to the literature on populist constitutionalism according to which the power that courts have gained in modern democracies, for example through judicial review, is undemocratic [@Tushnet.2000; @Waldron.2006; @Waldron.2021]. Instead, the @Tushnet.2000 argues that the people should have the power over the constitution. According to this ideological understanding of populism and constitutionalism, populists adapt the content of the constitution to the preferences of the current political majority [@Tushnet.2021].

An amended constitution might not reflect the pluralist elements of a liberal democratic constitution anymore. While liberal democratic constitutions focus on fundamental and human rights, separation of powers, and these days also on international integration, populist parties focus on the constitution as the embodiment of majoritarian preferences. "Constituent power, rather than being the power of the multitude, becomes the power of the majority."[@Blokker.2020d]

XXX GREECE

To translate the people's preferences into constitutional law, the constitution needs to be amended. Thus if populists behave this way once they gain power, we should observe an increase in constitutional changes once populists gain office.

> **Hypothesis 1:** The more populist a government, the higher the likelihood of constitutional amendments and replacements.

An additional implication of this approach is that these constitutional changes include the public in the amendment process. In order to meet the claim of general participation, from a democratic perspective, the design of the new constitution or constitutional amendments must also be inclusive [@LaTorre.2018]. Instead of an amendment proposal by the executive, the amendments should be developed, or at least discussed, in some kind of public forum. For example, the Bolivian constitution of 2009 was deliberated in a constitutional assembly and passed by a public referendum [@LaTorre.2020]. Unfortunately, we lack comparative data on how constitutional changes were drafted.

**The instrumental approach**

Other authors are more skeptical about the relationship between populism and constitutionalism. They argue that populists have a merely instrumental relationship to constitutionalism [@Muller.2017; @Mudde.2021; @CristobalRoviraKaltwasser.2013; @Scheppele.2019]. @Mudde.2021, for example, does not disagree with the theory of the ideological approach to populism and constitutionalism but argues that populism in practice does not necessarily equal populism in theory. Instead of bringing the constitution closer to the people, populists use it to their advantage. @Muller.2016 (p.62-62) writes: "when in power, populists tend to be much less skeptical about constitutionalism as a means of creating constraints on what they interpret to be the popular will". Populists only change the constitution if it hinders their interests, not to incorporate the majority's preferences.

This is closely related to the argument made by @Muller.2017b that populists are anti-pluralistic and regard themselves as the incarnation of the general will (p. 78). If populists regard themselves as the only true representatives of the people, direct democratic mechanisms become unnecessary. According to this approach, populists should try to consolidate their power - among other things by abolishing possible barriers that have been implemented in a constitution. However, this approach does not require frequent constitutional changes as the ideological approach does since the constitution does not need to be adapted to changes in public preferences.

The process of undermining democratic norms under the guise of constitutional, or legal, changes has been labeled as constitutional retrogression [@Huq.2018], autocratic legalism [@Scheppele.2018], or abusive constitutionalism [@Landau.2013]. Particularly when officeholders and incumbents aim to ensure that their state continues to appear democratic, law gains weight in the process of democratic regression [@Landau.2013]. In the case of an unconstitutional act such as a coup, it matters little whether the law is adapted. But, revolving around popular sovereignty, populists aim to maintain at least some democratic pretense. XXX executive aggrandizement

To test whether populists use constitutional reforms in this way, we must look beyond the frequency of constitutional changes. We must consider the effect of constitutional changes on the quality of liberal democratic institutions, such as the separation of powers. We know that populists in power have at least some negative effects on liberal democratic institutions [@RuthLovell.2022; @Huber.2017; @Kenny.2020; @Vittori.2022; @Houle.2018]. But, so far, we only have anecdotal evidence whether constitutional reforms are a mechanism used by populists to pursue their goals [@LaTorre.2020].

> **Hypothesis 2:** Constitutional Change under populist governments have a negative impact on the quality of polyarchy.
>
> **Hypothesis 3:** Constitutional Change under populist governments have a negative impact on the quality of liberal democracy.

But, as most scholars agree, the effect of populism and democracy are not solely negative [@Blokker.2019c; @Mudde.2013b; @LaTorre.2020]. Especially when taking into account different democratic dimensions -- such as participation, inclusion or representation, populists can have a positive impact on democratic quality. Many scholars argue that populism supports the idea of direct democracy to give people a direct channel for participation [@Mastropaolo.2021; @Gherghina.2021b; @Angelucci.2024; @Mudde.2013b]. The populist focus on popular sovereignty can increase participatory elements of democracy. Authors repeatedly refer to Latin American cases in which participatory elements have been strengthened. In the case of Bolivia, for example, such direct participation mechanisms were implemented through a constitutional reform in 2009 [@Mudde.2013b]. Particularly, in democracies that are characterized by a strong elite, and possibly even corruption, the populism can have a positive aspect on the inclusion of the people in democratic processes [@Mudde.2013b].

> **Hypothesis 4:** Constitutional Change under populist governments have a positive impact on the quality of participation.

Moreover, in Latin America, constitutional amendments stemming from populist actors have strengthened social rights [See @LaTorre.2020, p. 1464] and the inclusion of indigenous groups in the political process [@Mudde.2013b, p. 162]. Right-wing populists on the other hand are more likely to exclude groups, such as migrants, from civil society. Thus, we expect a significant interaction effect between ideology, populism and constitutional change.

> **Hypothesis 5a:** Constitutional Change under left-wing populist governments have a positive impact on the quality of civil society.
>
> **Hypothesis 5b:** Constitutional Change under populist governments have a negative impact on the quality of civil society.

A similar argument can be made with regard to the democratic dimension of egalitarianism. Left-wing populists parties often emphasize that all people need equal opportunities to participate in a democracy and to be heard [@RuthLovell.2022]. The dimension of egalitarianism also includes the idea that in order to reach equal opportunities, the state is supposed to redistribute wealth, a typical left-wing demand [@Hilgers.2013]. Thus, we expect a positive effect of constitutional change on egalitarian democracy under left-wing populists but a negative effect under right-wing populists.

> **Hypothesis 6a:** Constitutional Change under left-wing populist governments have a positive impact on the quality of egalitarianism.
>
> **Hypothesis 6b:** Constitutional Change under populist governments have a negative impact on the quality of egalitarianism.

# 3. Data and Empirical Strategy

To gain an overview of how populists in office use constitutional changes, we use data from the V-Dem [@vdemdata.2022], V-Party [@VParty.2022] and Comparative Constitutions Projects [@ComparativeConstitutions.2005]. We run fixed-effects models to estimate the effect of populists in power on constitutional changes and their effects.

To determine the populism score of a government, we draw on the populism index from the V-Party project [@VParty.2022], which ranks parties on a populism scale based on expert assessments of the rhetoric of party representatives regarding their anti-elitist attitudes and their reference to the people as a homogeneous group (`v2xpa_popul`). When a government consists of multiple parties, an average was taken of the populism scores of the parties involved, each weighted by their relative strength (measured by percentage of seats) within the governing coalition. This results in a populism index for governments between 0 (not populist) and 1 (populist).

We code whether a government is left- or right-wing based on the government parties' economic left-right scores [@VParty.2022]. Again, we calculate a weighted score that ranges from -3.43 (far-left) to 2.99 (far-right). Using an economic left-right scale omits other political dimensions which are important to understanding party competition in Europe [@Huber.2022]. In Latin America, party competition is mostly structured on the economic left-right axes [@MartinezGallardo.2023]. To be able to compare effects across continents, we use the economic left-right score. For European countries, we include a model in the appendix which uses the GAL-TAN score [@Jolly.2022].

In our first analysis, the occurance of a constitutional change is the dependent variable. The data comes from the Comparative Constitutions Project [@ComparativeConstitutions.2005]. We use the variable `evnt` that indicates whether a constitutional change, through amendment or replacement, occurred (0 -- no change, 1 -- change).

In the following models, the variable is used in a triple interaction effect to estimate the impact of constitutional changes under populist governments depending on their host-ideology on democratic quality. We measure the quality of democracy with the democracy indices from the Varieties of Democracies (V-Dem) project [@Vdem.2021]. Similar to @RuthLovell.2022, we use the different indices on participation, civil society, electoral, egalitarian, and liberal democracy to estimate the effect of changes under populist governments on different dimensions of democracy. Each index describes on a scale of 0 to 1 to what extent the ideals of the democratic dimension are implemented in the respective country-year observation.

A simple descriptive of what rights were changed when populist governments changed the constitution seems to confirm our expectations. We categorizes the content of constitutional amendments in executive power, right of the judiciary, as wells social and political rights.[^1]

[^1]: The strength of the executive is measured by an index developed by @Melton.2014. This measures whether the executive has the power to initiate legislation or constitutional amendments, issue decrees, declare a state of emergency, as well as enforce its power over other institutions through veto power, and have rights reviewed for constitutionality or dissolve parliament. The index of independent judiciary rights is based on @Melton.2014 and measures the number of constitutional norms that strengthen an independent judiciary (Included are the independence of the judiciary in the constitution, whether at least two actors are involved in the nomination and appointment of judges to the Constitutional Court, whether the dismissal of judges is severely restricted and limited only to serious misconduct or constitutional violations, and whether judges' salaries are protected. Instead of including lifetime appointments, we include whether the re-election of judges is excluded.) The index of political rights includes the guarantee of freedom of expression, as well as freedom of assembly, science, press, strike and trade union rights. Social rights include the guarantee of a certain standard of living, health protection at work, financial support, social security, and the right to a fair trial.

![Content changes of constitutional amendments by populist governments (government weighted populism score \> 0.5) that led to democratic regression.](results/graphs/rights_change.pdf){#fig-rightschange width="100%"}

If we now look at the content of populist (weighted populism score \> 0.5) we observe that there are multiple cases in which the rights of the executive were increased. At the same time, we also see an increase in social and political rights for citizens. However, the visual only includes 10 of the 84 countries that have experienced changes under populist governments (@tbl-populistchanges). In the other 74 cases, the rights indices did not capture any changes. We argue that we need to look beyond these salient cases, to get a better understanding of how populists use constitutions.

We restrict our analysis to 57 European and Latin American countries over the period 1991-2020. Country-year observations in which a regime is classified as a closed autocracy by V-Dem are excluded from our analysis, thus only the effect on already democratic states is analyzed. The the included time-frame ranges between 13 and 29 years for different countries. In total, our data panel consists of 1543 country-year observations. Of these, constitutional changes take place in 596 observations. However, constitutional events by governments with a high populism score of more than 0.5 are relatively rare: Only 90 observations meet this criterion. Table 1 provides an overview of these cases.

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-populistchanges
#| tbl-cap: Constitutional changes under populist governments (weighted populism score > 0.5)

# Read the .tex file content
tex_content <- readLines("results/tables/populistchanges.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")
```

In Europe, with the exception of Italy and Malta, constitutional changes by populist governments seem to be a rather Eastern European phenomenon. Also notable is that it is mainly right-wing populist governments that have been able to implement constitutional changes. A different picture emerges in Latin America. Here, constitutional changes have been implemented predominantly by left-wing populist governments.

Figure 1 shows the distributions of the populism index in Europe and Latin America. Populist governments are not exceptional in Latin America). In Europe, the distribution is clearly skewed to the right, with a majority of governments with few populists. The majority of European country-year observations have a relatively low populism score (less than 0.5).

![Distribution of the Populism Score of Governments in Europe and Latin America](results/graphs/histogram_populistgov.pdf){fig-align="center"}

To analyze whether constitutional changes by populist governments have an effect on democratic quality, we use a country fixed-effects panel model. How often constitutions are changed dependends on a country's rules and norms. Thus, we only use the within-country variance to estimate the effect of populists in office. We use the dependent variables one year ahead ("lead") for each of the democratic dimensions.[^2] A constitutional change in one year will probably not have an impact on the democracy index in the same year, but only with a slight delay.

[^2]: Models for different numbers of leads can be found in @tbl-leadlibdem to @tbl-leadegaldem .

# 4. Results

@tbl-resultschange shows the results for hypothesis 1. We do not find any evidence that populists in government are more likely to change the constitution. Neither left-wing nor right-wing populists governments implement constitutional changes more frequently than other governments (model 3). Neither is this the case if we look at the continents separately and use the GAL-TAN score as the left-right measurement for European parties (see appendix @tbl-resultschangeEU). We run an additional model that interacts the share of surplus seats a government has with its populism score to ensure that this result is not conditioned by the lack of a majority for some populist governments (@tbl-resultschange, model 4).

\blandscape

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-resultschange
#| tbl-cap: Regression results for likelihood of constitutional change

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchangelikelihood_evnt.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

\elandscape

The results shown in @tbl-results confirm our hypothesis that constitutional changes by left- and right-wing populist governments have different implications for democratic quality.[^3] For most democratic dimension we see a significant difference between left- and right wing populists in power.

[^3]: The results are robust for different leads (1-4), see appendix @tbl-leadlibdem to @tbl-egaldem.

Different to our expectations, we find evidence that populist governments use the constitutional changes to increase the quality of liberal democratic institutions (@tbl-libdem, model 4). Thus, we must reject our first hypothesis. The inclusion of the triple-interaction effect between populism, constitutional changes and host-ideology shows that this effect is driven by left-wing populist in power (@tbl-results, model 1 & @fig-interaction). We find similar results when we consider the quality of polyarchy ( @tbl-results, model 2, appendix @tbl-polyarchy , model 4 and @fig-interaction). These results are surprising given the general tenor that populists in office are a danger to the power separation of liberal democracies [@Huber.2017; @Muller.2016; @Mudde.2012b; @LaTorre.2020; @Kenny.2020; @Houle.2018]. Our results do not mean that this effect does not exist. But, the results imply that constitutional changes are not a main mechanism populists use to consolidate their power.

As expected, we find a significant positive effect of constitutional changes under left-wing populist governments on egalitarian democracy (@tbl-results, model 4 & @fig-interaction). Similarly, our hypothesis that constitutional changes under right-wing populist governments have a negative effect on egalitarian democracy is also confirmed. Indeed, egalitarian democracy is the only democratic dimension for which we find a significant effect of constitutional changes under right-wing populist government.

\blandscape

```{r}
#| output: asis
#| echo: false
#| warning: false
#| label: tbl-results
#| tbl-cap: Main Regression Models

# Read the .tex file content
tex_content <- readLines("results/tables/constitutionalchange_main.tex")

# Print the content as raw LaTeX
cat(tex_content, sep="\n")

```

\elandscape

Surprisingly, our hypotheses on participation and civil society do not hold. Neither the quality of participation nor civil society increases when populist governments change the constitution (appendix @tbl-partip & @tbl-cspart). This effect is independent of the governments ideological leaning (@tbl-results, models 3 & 5).

![Average marginal effect of constitutional change conditioned by government ideology and government populism score.](results/graphs/change_effect.pdf){#fig-interaction fig-align="center"}

Of course, left-wing populists in power in Europe have been rare to date. The positive effect of left-wing populists could be driven by the relatively young nature of Latin American democracies. If liberal democratic institutions are not only developing, populist parties in power might have a more positive impact on these institutions [@RuthLovell.2022]. We test this by replacing the ideology dummy with the democratic quality ahead of the respective country-year observation (lag of 2). @tbl-dynamic in the appendix shows that the triple-interaction effect is not significant in most of the models. Only in the case of the quality of civil society do we find that the effect differs depending on the quality ahead of the observation. But, the effect is different than expected: In countries that have had a lower quality of civil society, populists in government have an even stronger negative effect the former (see @fig-dynamic).

We measure ideology on an economic dimension to be able to compare the scale between continents. Our models that only include European parties and use their GAL-TAN score only show significant differences with regard to egalitarian democracy (appendix @tbl-resultschange_EU, model 4). If we use the GAL-TAN score, we do not find a significant negative effect on egalitarianism anymore. However, including only European states decreases the number of constitutional changes in our analysis to only 42, what makes it harder to find significant effects.

Our results are robust for different leads of the dependent variable (appendix @tbl-leadlibdem to @tbl-leadegaldem) and if we run jackknife models (appendix @fig-jackknife1, @fig-jackknife2, @fig-jackknifechange).

## DISCUSSION

To our knowledge, this is one of the first papers that evaluates the impact of constitutional changes under populist governments in a large-N analysis. Our results are not what one would expect: We do not find evidence that populists frequently adapt the constitution (to mirror the interest of the people), nor do we find significant negative effects of constitutional changes under populist governments on liberal democratic institutions.

Our research design has some limitations. We do not analyze the content of constitutional changes itself, we only use the change in democratic quality after a constitutional change. This leaves the possibility that populist governments implement the expected changes in the constitution, such as limitations on checks and balances or participation mechanisms, but that these do not have an effect on the quality of democracy. But, @fig-rightschange indicates that only in few cases the expected rights were actually implemented.

Both, our analysis as well as the description of what rights were changed under populist governments show that there is a heterogeneity between populist parties. Not all populists are the same. Our results show that left-wing populists have even had a positive effect on liberal democracy through their constitutional changes. Our results endorse the calls of scholars to regard populism as a heterogenous phenomenon that is not necessarily bad for all aspects of democracy [@Mudde.2012b; @Mudde.2013b; @Blokker.2019b].

Considering the relationship between constitutionalism and populism, we argue that our results have two main implications: Our understanding of constitutional changes under populist governments are too informed by few salient cases such as Hungary. Once we look at the bigger picture, we do not find the expected negative effect on liberal democratic institutions and polyarchy. Neither do we find the positive effect on participation that we have observed in some countries, such as Bolivia, on a large scale.

Our research further implies that we often overestimate the role that constitutions play for populists. We do not find any evidence for arguments that view populists as proponents of judicial populism [@Tushnet.2021]. Neither did populists adapt the constitution more often which could have shown a frequent adaptation to the public's interests, nor do they improve participation through their constitutional changes.

# 5. Conclusion

In this analysis, we focused on the relationship between populism in power and constitutional changes. We do not find any evidence for, what we call, the ideological approach to populism and constitutionalism. According to this approach populists aim to bring the constitution back to the people by frequently adapting it to the people's interests [@Tushnet.2021]. Our results do not show any evidence that populists in power adapt the constitution more often than other governments.

Once we turn to the effect of constitutional changes under populist governments, our results show that there are no clear correlations between a decrease of democratic quality after constitutional changes and populist governments. Thus, our results contrast with the thesis that the presence of populist parties in governments always has a negative effect on the quality of democracy. While constitutional changes under left-wing populist have a positive impact on polyarchy, liberal and egalitarian democratic quality, we find that right-wing populist parties have a significant negative impact on egalitarian democracy. However, we do not find that constitutional changes under populist governments significantly affect the quality of participation and civil society.

Our findings also confirm the warnings against drawing too quick conclusions about the relationship between populism and constitutional change. Various authors repeatedly point to the heterogeneity of populist parties and to the fact that populism can also have positive effects, for example on participation [@Blokker.2019c; @LaTorre.2020; @Mudde.2013b]. Our analysis also shows that conclusions about a clear relationship between populism, constitutional change and democratic regression are also not possible due to the rarity of constitutional reforms. In our Large-N study, it is clear that the Hungarian experience is not transferable to populist parties in Latin America.

### References

::: {#refs}
:::

{{< pagebreak >}}

 {{< include appendix.qmd >}} 
