---
title: "Constitutional Changes Under Populist Governments"
author:
  - name: Jasmin Sarah König^[University of Hamburg. Jasmin wurde durch die Deutsche Forschungsgemeinschaft (DFG) – GRK 2503 gefördert. Address for Correspondence jasmin.sarah.koenig@uni-hamburg.de]
    email: jasmin.sarah.koenig@uni-hamburg.de
    affiliation: University of Hamburg
  - name: Tilko Swalve^[Leibniz University Hannover]
---

The relationship between populism and constitutionalism has been in the centre of few but contentious debates. While some contend that populists seek to empower the people through constitutional reform [@Tushnet.2021; @Tushnet.2020; @Blokker.2019c], others posit that populists merely instrumentalize constitutional mechanisms to consolidate their power [@Mudde.2021; @CristobalRoviraKaltwasser.2013; @Muller.2017]. However, unlike the discourse surrounding populism and democracy, this debate often lacks concrete empirical evidence to substantiate its various claims. Our study aims to address this void by examining how populist administrations enact and leverage constitutional change in practice across 57 European and Latin American states.

In modern democracies, values and principles, such as the separation of powers or civil rights, are often implemented in a constitution [@Habermas.2001; @Whittington.2010]. The codification of such principles is supposed to constrain public authority and protect citizens from state power [@StoneSweet.2002]. However, it also constrains popular sovereignty by constraining the power of the majority, for example, through checks and balances or minority rights.

The populist ideology, on the other hand, revolves around popular sovereignty [@Abts.2007]. Populists vow to give the power back to the people. In populism, the outcome of politics not a restriction of the latter [@Blokker.2019c; @Mudde.2012]. Generally, one can distinguish two ways to think about the relationship between populism and constitutionalism. Some scholars argue that populism aims to democratize constitutions by increasing participation and adapting the contents of the constitution to the preferences of the majority [@Blokker.2019c; @Tushnet.2021]. We call this the ideological approach. Others argue that populism has an instrumental approach to constitutionalism. They contend that populists adapt the constitution in ways it consolidates their power [@Mudde.2021; @CristobalRoviraKaltwasser.2013; @Muller.2017].

We argue that, if populists are acting ideologically, we should observe an increase in constitutional changes under populist governments. But, if populists merely try to consolidate their power through constitutional changes, the constitution does not need to be changed more frequently. Few significant changes are sufficient to reach such a goal.

Frequently, the decrease of democratic quality through constitutional changes is linked with populists in power [@Scheppele.2019; see for example @Arato.2019; @Lacey.2019]. Thus, independent of the frequency of constitutional changes, the question arises whether constitutional changes have a different effect on democratic quality when populists are in power compared to other governments. The heterogeneity of populist parties makes it challenging to pin down their effects on democratic quality [@Mudde.2013b; @Mudde.2012; @Meny.2002b; @Huber.2017]. We argue that the ambiguity of populism is likely to be mirrored in the effects of their constitutional amendments on democratic quality. Depending on the democratic dimension under consideration, the effect of populist constitutional reform is likely to differ. Constitutional changes under left-wing populist governments are likely to differ from those under right-wing populist governments.

We test both arguments with a dataset combined from @RuthLovell.2022, the V-Dem [@Vdem.2021] and V-Party [@VParty.2022] as well as the Comparative Constitutions Project[@ComparativeConstitutions.2005]. We run fixed-effects panel models which do not show any evidence in favor of the ideological approach to populism and constitutionalism. Our results show that populists do not implement constitutional changes more frequently than other governments, independent of their ideology and government size. We show that the effect of constitutional changes under populist governments differs depending on whether they are left-, or right-leaning and the democratic dimension under consideration. While constitutional changes under left-wing populists have a significantly positive effect on liberal democracy, polyarchy, and egalitarianism, those under right-wing populists have a significantly negative effect on egalitarianism.

## Populism & Constitutionalism

The most commonly used definition of populism in Political Science describes it as a thin ideology [@Mudde.2004]. Populism usually emerges tied to a "denser" host ideology, for example fascism or socialism, and yet populism its own ideological core. The populist ideology is based on the majority principle, in which only the "will of the people" is to be implemented [@Abts.2007]. Any constrains on this process are understood as an obstruction of democracy [@Abts.2007; @Mazzoleni.2020].

In its endeavor to implement full popular sovereignty, populism divides society into an evil elite and the good people [@Mudde.2004]. According to populists, the people recognize what is best for them, an idea that is often mirrored in sayings such as the people having a *common sense* or *general will*. Whoever prevents this general will from being implemented is described as an enemy of the people.

Liberal democracy, on the other hand, must always balance institutionalization and popular sovereignty [@Canovan.1999]. This inherent incongruence, between popular sovereignty as the highest ideal and restriction on the same, is always part of liberal democracies. One attempt to institutionalize and safeguard democracy is to enshrine certain rights, institutions and values into a constitution [@Habermas.2001; @Whittington.2010]. In many liberal democracies separation of power as well as checks and balances are implemented in a constitution to prevent a centralization of power, or a tyranny of the majority [@Abts.2007].

These safeguards are in contradiction to the populist ideal of democracy. Different to liberal democracy, populism has a clear alignment between the two poles of institutionalization and popular sovereignty. The "will of the people" must be implemented as quickly as possible and without obstacles. Institutions and norms serve the sole purpose of supporting this process, but must never hinder it [@Canovan.1999]. While liberal democratic constitutions focus on fundamental and human rights, separation of powers, and these days also on international integration, populist parties focus on the constitution as the embodiment of majoritarian preferences. "Constituent power, rather than being the power of the multitude, becomes the power of the majority."[@Blokker.2020d]

The discrepancy between populism and liberal democracy is also mirrored in the populist understanding of the law. Populism is based on the supremacy of the political [@Mudde.2012; @Blokker.2019d]. Law can accordingly only express the outcome of political processes, but can never justify their restriction. The claim of a neutral law that stands above the political process is not recognized by populists. Instead, law is seen as a purely political medium. The constitution is supposed to reflect the will of the majority of the people and is thus not seen as a firmly established institution that rarely changes but as a living document that is purely political [@Blokker.2020d; @Mazzoleni.2020]. Populists do not strive to abolish constitutions, but to re-politicize them in the sense of the alleged "will of the people" [@Mazzoleni.2020; @Muller.2017b].

**The ideological approach**

However, scholars are divided about what this implies for constitutionalism once populists are in power. According to the first approach, populists act purely ideological once in office and aim to give the power back to the people [@Tushnet.2021; @Blokker.2019d]. In order to always reflect the will of the majority, constitutions should be easy and quick to change according to the populist ideal [@Fabbrizi.2020]. With this understanding of the constitution, the distinction between ordinary and constitutional law is abolished [@Blokker.2020d]. If constitutional law is no longer seen as a guideline in the political process, but only as an expression of the political, it loses its elevated and particularly safeguarded position.

This understanding of populism and constitutionalism is closely related to the literature on populist constitutionalism according to which the power that courts have gained in modern democracies, for example through judicial review, is undemocratic [@Tushnet.2000; @Waldron.2006; @Waldron.2021]. Instead, @Tushnet.2000 argues that instead of the courts, the people should decide about the contents and interpretation of the constitution. The ideological approach views populists as populist constitutionalists [@Tushnet.2021]. According to this idea, populists aim to adapt the content of the constitution to the preferences of the current political majority and give the power over the constitution back to the people [@Tushnet.2021].

An additional implication of this approach is that these constitutional changes include the public in the amendment process. In order to meet the claim that the people have the power over the constitution, the design process of the new constitution or constitutional amendments must also be inclusive [@LaTorre.2018]. Instead of an amendment proposal by the executive, the amendments should be developed, or at least discussed, in some kind of public forum. For example, the Bolivian constitution of 2009 was deliberated in a constitutional assembly and passed by a public referendum [@LaTorre.2020]. Unfortunately, we lack comparative data on how constitutional changes were drafted and can not consider this implication in our analysis.

**The instrumental approach**

Other authors are more skeptical about the relationship between populism and constitutionalism. They argue that populists have a merely instrumental relationship to constitutionalism [@Muller.2017; @Mudde.2021; @CristobalRoviraKaltwasser.2013; @Scheppele.2019]. These authors do not necessarily disagree with the theory of the ideological approach to populism and constitutionalism [see for example @Mudde.2021]. However, they argue that populism in practice does not equal populism in theory. @Muller.2016 (p. 62-63) writes: "when in power, populists tend to be much less skeptical about constitutionalism as a means of creating constraints on what they interpret to be the popular will". Instead of bringing the constitution closer to the people, populists use it to their advantage. According to this approach, populists change the constitution if it hinders their interests, not to incorporate the majority's preferences.

This is closely related to the argument made by @Muller.2017b that populists are anti-pluralistic and regard themselves as the incarnation of the general will. If populists think of themselves as the only true representatives of the people, direct democratic mechanisms or changes in government become unnecessary. Instead, as soon has they had to leave office, the people's will would not be implemented anymore. According to this approach, populists try to consolidate their power - among other things by abolishing possible barriers that have been implemented in a constitution.

We argue that, if populists aim to consolidate their power through constitutional amendments because they regard themselves as the incarnation or only representation of the people, few - though significant - changes of the consitution should be sufficient. But, if populists actually aim to translate the people's preferences into constitutional law, they need to amend the constitution more frequently than other governments.

> **Hypothesis 1:** The more populist a government, the higher the likelihood of constitutional amendments and replacements.

### The Effect of Constitutional Changes on Democratic Quality

Another way to think about constitutional changes under populist governments is to consider their effects on democratic quality [@Haas.2022; @Konig.2023b]. The relationship between democracy and populism is controversial [@Canovan.1999; @Meny.2002b; @Mouffe.2005; @Laclau.2005; @Abts.2007; @Mudde.2012b; @Konig.2023b]. While many authors agree that populism and liberal democracy are at least partially in conflict, they frequently also point out that populism may well have a corrective effect on the responsiveness of or participation in democracies [@LaTorre.2020; @Mudde.2012b; @Canovan.1999]. The relationship becomes even more complicated because populism is only a thin ideology [@Mudde.2012]. The relationship is likely to differ between different host-ideologies [@Huber.2017] and democratic dimensions [@RuthLovell.2022]. We consider both aspects by discussing the different effects of populism on polyarchy, civil society, participation, liberal and egalitarian democracy. In cases where we expect the effect to differ based on ideology, we discuss the possible different effects between left- and right-wing populists.

We first consider the effects of constitutional changes under populist governments on liberal democratic quality. Populism is in conflict in liberal democratic institutions that constrain the power of the majority [@Abts.2007]. Empirical evidence corroborates this claim. Populists in power have at least some negative effects on liberal democratic institutions though the extent of it differs between studies [@RuthLovell.2022; @Huber.2017; @Kenny.2020; @Vittori.2022; @Houle.2018].

Particularly when officeholders and incumbents aim to ensure that their state continues to appear democratic - as populists with their claim to be radical democratic must - legal changes are an important step in consolidating power [@Landau.2013]. A re-written constitution can legitimize the executive's grip on power. This process of undermining democratic norms under the guise of constitutional, or legal, changes has been labeled as constitutional retrogression [@Huq.2018], autocratic legalism [@Scheppele.2018], or abusive constitutionalism [@Landau.2013]. But, so far, we only have anecdotal evidence whether constitutional reforms are a mechanism used by populists to pursue their goals [@LaTorre.2020; @Muller.2017; @AydinCakir.2023; @Arato.2019; @Scheppele.2019]. If populists use constitutional changes in such an instrumental way, we should observe a decrease in the quality of liberal democracy and polyarchy after such changes.

> **Hypothesis 2:** Constitutional Change under populist governments has a negative impact on the quality of liberal democracy.
>
> **Hypothesis 3:** Constitutional Change under populist governments has a negative impact on the quality of polyarchy.

But, as most scholars agree, the effects of populism on democracy are not solely negative [@Blokker.2019c; @Mudde.2013b; @LaTorre.2020]. Especially when taking into account different democratic dimensions, such as participation, inclusion or representation, populists can have a positive impact on democratic quality. The populist focus on popular sovereignty can increase participatory elements of democracy. Many scholars argue that populism supports the idea of direct democracy to give people a more direct channel for participation [@Mastropaolo.2021; @Gherghina.2021b; @Angelucci.2024; @Mudde.2013b]. Particularly, in democracies that are characterized by a strong elite, and possibly even corruption, populism can have a positive effect on the inclusion of the people in democratic processes [@Mudde.2013b]. This can be observed in Latin American cases in which participatory elements have been strengthened through constitutional reform. In the case of Bolivia, for example, such direct participation mechanisms were implemented through a constitutional reform in 2009 [@Mudde.2013b]. We expect that the focus on popular sovereignty leads to an increase in the quality of participation through constitutional changes under populist governments.

> **Hypothesis 4:** Constitutional Change under populist governments has positive impact on the quality of participation.

Depending on its host-ideology, populism can take an inclusive and an exclusive form [@Mudde.2013b]. Left-wing populists, on the one hand, often aim to incorporate neglected groups within society. In Latin America, constitutional amendments stemming from populist actors have strengthened social rights [@LaTorre.2020] and the inclusion of indigenous groups in the political process [@Mudde.2013b]. Right-wing populists on the other hand are more likely to exclude groups, such as migrants, from civil society. Thus, we expect a significant interaction effect between ideology, populism and constitutional change when we measure the effect of populism on the quality of civil society.

> **Hypothesis 5a:** Constitutional Change under left-wing populist governments has a positive impact on the quality of civil society.
>
> **Hypothesis 5b:** Constitutional Change under populist governments has a negative impact on the quality of civil society.

A similar argument can be made with regard to the democratic dimension of egalitarianism. Left-wing populist parties often emphasize that all people need equal opportunities to participate in a democracy and to be heard [@RuthLovell.2022]. The dimension of egalitarianism also includes the idea that in order to reach equal opportunities the state is supposed to redistribute wealth, a typical left-wing demand [@Hilgers.2013]. Thus, we expect a positive effect of constitutional change on egalitarian democracy under left-wing populists but a negative effect under right-wing populists.

> **Hypothesis 6a:** Constitutional Change under left-wing populist governments has a positive impact on the quality of egalitarianism.
>
> **Hypothesis 6b:** Constitutional Change under populist governments has a negative impact on the quality of egalitarianism.

## Data & Research Design

### Data

To gain an overview of how populists in office use constitutional changes, we use data from the V-Dem [@vdemdata.2022], V-Party [@VParty.2022] and Comparative Constitutions Projects [@ComparativeConstitutions.2005]. We restrict our analysis to 57 European and Latin American countries over the period 1991-2020. Country-year observations in which a regime is classified as a closed autocracy by V-Dem are excluded from our analysis, thus only constitutional changes in already democratic states are analyzed. Due to missing data, the included time-frame ranges between 13 and 29 years for different countries. In total, our data panel consists of 1 543 country-year observations. Of these, constitutional changes take place in 596 observations. However, constitutional events by governments with a high populism score of more than 0.5 are relatively rare: Only 84 observations meet this criterion.[^1]

[^1]: @tbl-populistchanges in the appendix provides an overview of these cases

The inclusion of both, European and Latin American states, allows us to use the variance of left- and right-wing populists in power. In Europe it is mainly right-wing populist governments that have been able to implement constitutional changes. A different picture emerges in Latin America. Here, constitutional changes have been implemented predominantly by left-wing populist governments.

### Research Design

To analyze whether constitutional changes by populist governments have an effect on democratic quality, we use country fixed-effects panel models. To account for heterogeneity, we use robust standard errors. chrome://settings/search

How often constitutions are changed depends on a country's rules and norms. Thus, we only use the within-country variance to estimate the effect of populists in office. A constitutional change in one year will probably not have an impact on the democracy index in the same year, but only with a slight delay. We use the dependent variables one year ahead for each of the democratic dimensions.[^2]

[^2]: Models for different numbers of leads can be found in @tbl-leadlibdem to @tbl-leadegaldem .

#### Operationalization

For our independent variable, we draw on the populism index from the V-Party project [@VParty.2020, `v2xpa_popul`].[^3] The score ranks parties on a populism scale based on expert assessments of the rhetoric of party representatives regarding anti-elitist attitudes and whether they reference the people as a homogeneous group. When a government consists of multiple parties, we weighed each parties populism score by their relative strength[^4] within the governing coalition. This results in a populism index for governments between 0 (not populist) and 1 (populist).

[^3]: The populism score from the V-Party dataset comes with some ambiguities. It allows researchers to conduct long-term cross-country analyses, which so far was only possible with a large amount of manual coding [@Huber.2017; @Kenny.2020; @RuthLovell.2022; @Ruth.2018]. But, the retrospective expert coding of parties can lead to biased assessments [@Little.2024]. Further, the score does not include any item on the distribution of power in democracies, a crucial driver of populists' stance on constitutionalism.

[^4]: Measured by percentage of seats

@fig-populism shows the distributions of the populism index in Europe and Latin America. Populist governments are not exceptional in Latin America. In Europe, so far, there were only few governments that were highly populist - the distribution is clearly skewed to the right. The majority of European country-year observations have a relatively low populism score (less than 0.5).

![Distribution of the Populism Score of Governments in Europe and Latin America](results/graphs/histogram_populistgov.pdf){#fig-populism alt="Distribution of the Populism Score of Governments in Europe and Latin America" fig-align="center"}

We code whether a government is left- or right-wing based on the parties' economic left-right scores [@VParty.2022]. Again, we calculate a weighted score that ranges from -3.43 (far-left) to 2.99 (far-right). We use the economic left-right score to be able to compare effects across continents. While this omits other political dimensions which are important to understanding party competition in Europe [@Huber.2022], party competition in Latin America is mostly structured on the economic left-right axes [@MartinezGallardo.2023]. To consider the two-dimensional policy space of European party systems, we include models for European countries in the appendix which use the GAL-TAN score [@Jolly.2022].

In our first analysis, the occurrence of constitutional change is the dependent variable. The data comes from the Comparative Constitutions Project [@ComparativeConstitutions.2005, `evnt`]. We use the variable that indicates whether a constitutional change, through amendment or replacement, occurred (0 -- no change, 1 -- change).

In the following models, we estimate the effect of constitutional changes under populist governments on the quality of democracy. To consider the governments' ideology, we use a triple interaction effect. We interact the dummy on constitutional change with the government left-right score with the government populism scoree.

We measure the quality of democracy with the democracy indices from the Varieties of Democracies (V-Dem) project [@Vdem.2021]. Similar to @RuthLovell.2022, we use the different indices on participation, civil society, electoral, egalitarian, and liberal democracy to estimate the effect of constitutional changes under populist governments on different dimensions of democracy. Each index describes on a scale of 0 to 1 to what extent the ideals of the democratic dimension are implemented in the respective country-year observation.

Since constitutional changes often need a certain majority in parliament, we control for the share of surplus seats a government has. The more parties are part of a coalition, the harder it gets to agree on a reform. We include a coalition dummy as a control.

## Results

We do not find support for the ideological, nor the instrumental approach to populism and constitutionalism. Instead, we find a result neither of both approaches would have predicted. Left-wing populists improved the quality of liberal democracy and polyarchy through constitutional changes. However, we do find the expected differences between constitutional changes under left- and right-wing populists. While left-wing populists use these for a positive impact on the democratic dimensions of liberal democracy, polyarchy and egalitarianism, constitutional changes under right-wing populists have only had a negative impact on the quality of egalitarianism and no effect on all other dimensions of democracy.

### Frequency of Constitutional Changes

@tbl-resultschange shows the results for hypothesis 1. We do not find any evidence that populists in government are more likely to change the constitution. Neither left-wing nor right-wing populists governments implement constitutional changes more frequently than other governments (model 3). Neither is this the case if we look at the continents separately and use the GAL-TAN score as the left-right measurement for European parties (see appendix @tbl-resultschange_EU).

We run an additional model that interacts the share of surplus seats a government has with its populism score to ensure that this result is not conditioned by the lack of a majority for some populist governments (@tbl-resultschange, model 4). However, the model returns similar results what implies that the lack of constitutional changes under populist governments does not arise from a lack of necessary majorities.

### Democratic Quality

The results shown in @tbl-results confirm our hypothesis that constitutional changes by left- and right-wing populist governments have different implications for democratic quality.[^5] For most democratic dimension we see a significant difference between left- and right wing populists in power.

[^5]: Our results are robust for different leads of the dependent variable (appendix @tbl-leadlibdem to @tbl-leadegaldem) and if we run jackknife models (appendix @fig-jackknife1, @fig-jackknife2, @fig-jackknifechange).

\blandscape

```{r} #| output: asis #| echo: false #| warning: false #| label: tbl-resultschange #| tbl-cap: Regression results for likelihood of constitutional change  # Read the .tex file content tex_content <- readLines("results/tables/constitutionalchangelikelihood_evnt.tex")  # Print the content as raw LaTeX cat(tex_content, sep="\n")}
```

\elandscape

Our results do not support our second or third hypotheses on liberal democracy and polyarchy. Indeed, we find evidence that populist governments have used constitutional changes to increase the quality of liberal democratic institutions (@tbl-resultsinteraction, model 1). The inclusion of the triple-interaction effect between populism, constitutional changes and host-ideology shows that this effect is driven by left-wing populists in power (@tbl-results, model 1 & @fig-interaction). We find similar results when we consider the quality of polyarchy with regard to hypothesis 3 ( @tbl-resultsinteraction, model 2 and @tbl-results, model 2). Left-wing populists have improved polyarchy through constitutional changes, while right-wing populists did not have any signficiant effect on the quality of polyarchy (@fig-interaction).

These results are surprising given the general tenor that populists in office are a danger to the limits of executive power in liberal democracies [@Huber.2017; @Muller.2016; @Mudde.2012b; @LaTorre.2020; @Kenny.2020; @Houle.2018]. Our results do not mean that this effect does not exist. But, the results imply that constitutional changes are not a main mechanism populists use to undermine liberal democratic institutions.

As expected, we find a significant positive effect of constitutional changes under left-wing populist governments on egalitarian democracy (@tbl-results, model 4 & @fig-interaction). Similarly, our hypothesis that constitutional changes under right-wing populist governments have a negative effect on egalitarian democracy is also confirmed. Indeed, egalitarian democracy is the only democratic dimension for which we find a significant effect of constitutional changes under right-wing populist governments.[^6]

[^6]: We measure ideology on an economic dimension to be able to compare the scale between continents. Our models that only include European parties and use their GAL-TAN score only show significant differences with regard to egalitarian democracy (appendix @tbl-resultschange_EU, model 4). If we use the GAL-TAN score, we do not find a significant negative effect on egalitarianism anymore. However, including only European states decreases the number of constitutional changes in our analysis to only 42, what makes it harder to find significant effects.

Surprisingly, our hypotheses on participation and civil society do not hold. Neither the quality of participation nor civil society increases when populist governments change the constitution (appendix @tbl-resultsinteraction, model 3 & 5). This effect is independent of a government's ideological leaning (@tbl-results, models 3 & 5).

\blandscape

```{r} #| output: asis #| echo: false #| warning: false #| label: tbl-resultsinteraction #| tbl-cap: Fixed-Effects Panel Models with Interaction between Constitutional Change and Populism.  # Read the .tex file content tex_content <- readLines("results/tables/constitutionalchange_interaction.tex")  # Print the content as raw LaTeX cat(tex_content, sep="\n")}
```

\renewcommand{\arraystretch}{0.8}

```{r} #| output: asis #| echo: false #| warning: false #| label: tbl-results #| tbl-cap: Fixed-Effects Panel Models with Interaction between Constitutional Change, Populism and Ideology.  # Read the .tex file content tex_content <- readLines("results/tables/constitutionalchange_main.tex")  # Print the content as raw LaTeX cat(tex_content, sep="\n")}
```

\elandscape

![Average marginal effect of constitutional change conditioned by government ideology and government populism score.](results/graphs/change_effect.pdf){#fig-interaction alt="Average marginal effect of constitutional change conditioned by government ideology and government populism score." fig-align="center"}

### Alternative Explanations

**Dynamic Models**

As argued before, the effect of populists in power on democratic quality can also depend on how a democracy has developed before populists came into power or implemented constitutional changes. In a democracy that is very elitist, populists can have a more positive impact on participation than in a democracy where citizens already have a wide range of possibilities to engage [@Mudde.2012; @Mudde.2017b]. In the case of liberal democracy and polyarchy, younger institutions are more volatile to threats since they still need to build up there legitimacy [@RuthLovell.2022; @Gibler.2011]

We test whether this has an impact on our results by replacing the ideology dummy with the democratic quality ahead of the respective country-year observation. We now interact the populism score, constitutional changes and the lag (2) of the dependent democracy dimension. @tbl-dynamic in the appendix shows that the triple-interaction effect is not significant in most models. Only in the case of the quality of civil society do we find that the effect differs depending on the quality before the observation. In countries that have had a lower quality of civil society, populists in government have a stronger negative effect on the former (see appendix @fig-dynamic). However, the robustness check does not show any results that imply that the former quality of democracy drives the significant results in our main models on the dimensions of liberal democracy, polyarchy and egalitarianism.

**Measuring Changes in Constitutional Rights**

Our research design has some limitations. We do not analyze the content of constitutional changes itself, we only use the change in democratic quality after a constitutional change. This leaves the possibility that populist governments implement the expected changes in the constitution, such as limitations on checks and balances or participation mechanisms, but that these do not have an effect on the quality of democracy. We try to answer this data by building indices on rights for the executive and the judiciary, political and social rights based on the CCP data.[^7]

[^7]: The strength of the executive is measured by an index developed by @Melton.2014. This measures whether the executive has the power to initiate legislation or constitutional amendments, issue decrees, declare a state of emergency, as well as enforce its power over other institutions through veto power, and have rights reviewed for constitutionality or dissolve parliament. The index of independent judiciary rights is based on @Melton.2014 and measures the number of constitutional norms that strengthen an independent judiciary (Included are the independence of the judiciary in the constitution, whether at least two actors are involved in the nomination and appointment of judges to the Constitutional Court, whether the dismissal of judges is severely restricted and limited only to serious misconduct or constitutional violations, and whether judges' salaries are protected. Instead of including lifetime appointments, we include whether the re-election of judges is excluded.) The index of political rights includes the guarantee of freedom of expression, as well as freedom of assembly, science, press, strike and trade union rights. Social rights include the guarantee of a certain standard of living, health protection at work, financial support, social security, and the right to a fair trial.

The indices on executive and judicial power only show any changes in 6 out of 84 cases with a populism score \> 0.5. In these cases, the data shows an increase in executive rights and a decrease in rights for the judiciary (see @fig-rightschange). However, a very large majority of constitutional changes under populist governments did not aggrandize power.[^8]

[^8]: We test whether this relationship is significant in a regression model shown in the appendix, @tbl-executive. The results show that even if a populist government has a large share of surplus seats, there is no increased likelihood of an increase in rights for the executive (see appendix @fig-executive).

![Content of constitutional changes by populist governments (government weighted populism score \> 0.5. Only cases in which at least one of the described indices changed are shown.](results/graphs/rights_change.pdf){#fig-rightschange alt="Content of constitutional changes by populist governments (government weighted populism score > 0.5. Only cases in which at least one of the described indices changed are shown." fig-scap="Content of constitutional changes by populist governments (government weighted populism score > 0.5." width="100%"}

Considering social and political rights, we see a mixed picture. Some populist governments seem to have increased political and social rights, while in other cases particularly social rights were decreased (see @fig-rightschange). But, the indices only show differences for 9 out of the 84 cases with a populism score \> 0.5. While the data corroborates the argument that not all populist parties in power act in the same way, the small number of cases in which we can measure a change in written constitutional rights, does not allow for any inferences to all populist governments.

## Conclusion

To our knowledge, this is one of the first papers that evaluates the impact of constitutional changes under populist governments in a large-N analysis. Our findings confirm the warnings against drawing too quick conclusions about the relationship between populism and constitutional change [@Tushnet.2021; @Blokker.2019b]. We can neither find support for the arguments that populists frequently adapt the constitution to the will of the majority[@Tushnet.2021], neither do we observe that populists use such changes to aggrandize power [@Muller.2017; @Mudde.2021].

Instead, our results show that left-wing populists have even had a positive effect on liberal democracy, polyarchy and egalitarianism through their constitutional changes. Constitutional reforms under right-wing populist governments only had a negative effect on the quality of egalitarianism. However, we do not find that constitutional changes under populist governments significantly affect the quality of participation and civil society.

On the one hand, our results are in contrast with the thesis that populists in governments use constitutional changes to undermine liberal democracy [@Muller.2017; @Mudde.2021; @Scheppele.2019; @Arato.2019]. On the other hand, they corroborate the calls of scholars to regard populism as a heterogeneous phenomenon that is not necessarily bad for all aspects of democracy [@Mudde.2012b; @Mudde.2013b; @Blokker.2019b]. Both, our analysis as well as the description of what rights were changed under populist governments show that there is a heterogeneity between populist parties. Not all populists are the same.

Considering the relationship between constitutionalism and populism, we argue that our results have two main implications: Our understanding of constitutional changes under populist governments are too informed by few salient cases, such as Hungary. Once we look at the bigger picture, we do not find the expected negative effect on liberal democratic institutions and polyarchy. Neither do we find the positive effect on participation that we have observed in some countries, such as Bolivia, on a large scale.

Our research further implies that we often overestimate the role that constitutions play for populists [@Mudde.2021; @Blokker.2019e; @Muller.2017]. We do not find any evidence for arguments that view populists as proponents of judicial populism [@Tushnet.2021; @Blokker.2019e]. Neither did populists adapt the constitution more often which could have shown a frequent adaptation to the public's interests, nor do they improve participation through their constitutional changes.

Future research should consider in more detail how constitutional texts are changed and how populist voters view such changes. Further, the support for citizens for such changes is crucial in constitutional democracies [@Engst.2023]. But, so far, we know very little about how voters perceive constitutional changes under populist governments [@Magalhaes.2023]. This paper provides some evidence on how populist parties use constitutional changes once they are in power. Our approach to compare a 57 countries across two continents shows that we need to look beyond well known cases because populists use constitutional changes in variety of ways. We hope that others follow and more research on the complex relationship between populism and constitutionalism will follow.
